---
title: "HDAC1 and HDAC2 Intestinal Immunity - scRNAseq Quality control"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Quality control of scRNAseq of mouse intestinal CD4+ and CD8+ T cells

Here we will do the initial data quality control and demultiplexing of scRNAseq data

- Input: Raw data  

- Approach: Merge Experiments 1 and 2; Filter nFeature, NCount and MT %; Demultiplexing of Hashtag Oligos and CITEseq Oligcos; Filter doublets;  

- Output: QC Plots; Filtered Obj


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE)

```

## Libraries
```{r libraries, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(patchwork)
library(ggplot2)
```

## Dir
```{r parameters}
# set up input data dir
dir_raw <- here("Data", "Raw")
dir_processed <- here("Data", "Processed", "01_QualityControl")

# create if doesn't exist
dirs <- c(dir_processed)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)

```

## Read data and create Seurat objects

Steps in this loop:  

1. Read in datasets from Experiment 1 and Experiment 2

2. Create Seurat object for each

3. Add the hashtag oligos (HTO) data to the seurat object as an extra assay

4. Add the hashtag CITE-seq data to the seurat object as an extra assay

5. Create new mitochondrial percent and fraction to metadata

6. Get unique barcodes per experiment

7. Assign Seurat objects

8. Merge both replicates into one Seurat obj

9. Clean env

```{r read in data, warning=FALSE}

for (file in list.files(dir_raw)){ # For each experiment 1 and 2:
  
  # 1. Read in data and create Seurat object with counts  
  writeLines(paste("\nName:", file))
  
  seurat_data <- Read10X(data.dir = file.path(dir_raw, file))
  
  # 2. Create the Seurat object
  seurat_obj <- CreateSeuratObject(counts = seurat_data$`Gene Expression`, 
                                   min.features = 100,
                                   min.cells = 3, 
                                   project = file) 
  
  # 3. Add hashtag oligo data to object as extra assay
  seurat_obj[["HTO"]] <- CreateAssayObject(seurat_data$`Multiplexing Capture`[,colnames(seurat_obj)],
                                           min.cells = 1000) # already selects only those cells that passed the min.features threshold defined above
  writeLines(paste("Used hashtag oligos:"))
  print(rownames(seurat_obj@assays$HTO))
  
  # 4. Add hashtag CITE-seq data to object as extra assay
  seurat_obj[["CITE"]] <- CreateAssayObject(seurat_data$`Antibody Capture`[,colnames(seurat_obj)],
                                            min.cells = 1000) # selects only cells that passed min.features threshold defined before
  writeLines(paste("Used CITE-seq antibody targets:"))
  print(rownames(seurat_obj@assays$CITE))
  
  # 5. Add mitochondrial percent and fractionto object's metadata
  seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^mt-")
  seurat_obj[["fraction.mt"]] <- seurat_obj[["percent.mt"]]/100
  
  # 6. Prints number of unique cellular barcodes per experiment
  writeLines(paste("Information on Seurat object: "))
  print(seurat_obj)
  
  # 7. Assign Seurat object to folder name and remove old files
  assign(file, seurat_obj)
  rm(list=c("seurat_data", "seurat_obj"))
  gc() # clean environment
}


# 8. Merge the (biological) replicate experiments
mergedSeurat <- merge(x=Experiment1,
                      y=Experiment2,
                      add.cell.ids = c("Exp1", "Exp2")) #adds exp ids to barcodes

# add experiment information to metadata
mergedSeurat$Experiment <- mergedSeurat$orig.ident

# View(mergedSeurat@meta.data)

# 9. Remove unneeded data
rm(list=c("Experiment1", "Experiment2", "file"))
```

## Object overview

```{r object overview}
## Print the object
writeLines("Merged Seurat object:")
mergedSeurat
# An object of class Seurat 
# 20938 features across 80292 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: counts.Experiment1, counts.Experiment2
#  2 other assays present: HTO, CITE


## Show summary 
writeLines("Number of droplets per experiment:")
mergedSeurat$Experiment %>% table()
# Experiment1 Experiment2 
#       25690       54602 


## Plot QC metrics
VlnPlot(mergedSeurat,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3,
        pt.size = 0.01,
        alpha=0)

## Calculate median mitochondrial count percent
tapply(mergedSeurat$percent.mt, INDEX=mergedSeurat$Experiment, FUN="median")
# Experiment1 Experiment2 
#    3.115706    1.428571 

```

## Plots overview
```{r plots overview}
## Plot number of features per experiment
# how many genes detected
xlim_upper<- quantile(mergedSeurat@meta.data$nFeature_RNA, probs=0.9)

ggplot(mergedSeurat@meta.data) +
  geom_density(aes(x=nFeature_RNA, col=orig.ident)) + xlim(0, xlim_upper)

mergedSeurat@assays$RNA$counts.Experiment1 %>% colSums() %>% summary()
 # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 #  102.0   161.0   233.0   512.3   440.0 24700.0 
mergedSeurat@assays$RNA$counts.Experiment2 %>% colSums() %>% summary()
 # Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 #  105.0   138.0   162.0   974.6   513.0 34637.0 


## Plot rank vs nr Genes
# x: rank of each cell by y: total RNA counts
# identify true cells vs background noise (knee drop in plot)
mergedSeurat@meta.data$Rank_nCount_RNA <-  rev(1:length(mergedSeurat@meta.data$orig.ident))[rank(mergedSeurat@meta.data$nCount_RNA)]

ggplot(mergedSeurat@meta.data) +
  geom_point(aes(x=Rank_nCount_RNA, y=log10(nCount_RNA)), color="blue")

## Plot number of counts vs number of features, color by fraction of mitochondrial read counts
# high mito count in the bottom end of plot, specially in exp 2
ggplot(mergedSeurat@meta.data) +
  geom_point(aes(y=nFeature_RNA, x=nCount_RNA, col=fraction.mt)) +
  scale_color_continuous(type="viridis") + 
  facet_wrap(~orig.ident) +
  theme_bw()

```

## Filtering 

Filtering based on number of features and mito content

```{r filtering}
## Perform filtering
mergedSeurat<- subset(mergedSeurat,
                      subset = nFeature_RNA > 350 & nFeature_RNA < 4000 
                      & nCount_RNA > 500 
                      & percent.mt < 5)

mergedSeurat
# An object of class Seurat 
# 20938 features across 15536 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: counts.Experiment1, counts.Experiment2
#  2 other assays present: HTO, CITE

mergedSeurat$orig.ident %>% table()
# Experiment1 Experiment2 
#        4744       10792 


## Plot QC metrics after filtering
VlnPlot(mergedSeurat,
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
        ncol = 3,
        pt.size = 0.01,
        alpha=0)
```

## Demultiplex Hashtag oligos

Hashtag Oligos (HTOs): short DNA-barcoded antibodies, for sample multiplexing. Here we will demultiplex it by HTOs, to separate samples within each experiment. 

```{r demultiplex hashtag oligos}

## Subset for each individual experiment
Experiment1 <- subset(mergedSeurat, subset = Experiment == "Experiment1")
Experiment2 <- subset(mergedSeurat, subset = Experiment == "Experiment2")


## Normalize within each Experiment
Experiment1 <- NormalizeData(Experiment1, assay = "HTO", normalization.method = "CLR")
Experiment2 <- NormalizeData(Experiment2, assay = "HTO", normalization.method = "CLR")


## Demultiplex using HTODemux function
Experiment1 <- HTODemux(Experiment1, assay = "HTO", positive.quantile = 0.99)
# Cutoff for WT-SI-IEL : 17 reads
# Cutoff for KO-SI-IEL : 63 reads
# Cutoff for WT-SI-LPL : 6 reads
# Cutoff for KO-SI-LPL : 10 reads
# Cutoff for WT-colon : 7 reads
# Cutoff for KO-colon : 4 reads

Experiment2 <- HTODemux(Experiment2, assay = "HTO", positive.quantile = 0.99)
# Cutoff for WT-SI-IEL : 285 reads
# Cutoff for KO-SI-IEL : 13 reads
# Cutoff for WT-SI-LPL : 4 reads
# Cutoff for KO-SI-LPL : 7 reads
# Cutoff for WT-colon : 7 reads
# Cutoff for KO-colon : 4 reads


## Merge the (biological) replicate experiments again
# don't add exp1_ info again to avoid repetition. It merges by cell names.
# with HTO info in metadata
mergedSeurat <- merge(x=Experiment1,
                      y=Experiment2)

Idents(mergedSeurat) <- "HTO_maxID"

## HTO cols present
colnames(mergedSeurat@meta.data)
# [1] "orig.ident"                "nCount_RNA"               
#  [3] "nFeature_RNA"              "nCount_HTO"               
#  [5] "nFeature_HTO"              "nCount_CITE"              
#  [7] "nFeature_CITE"             "percent.mt"               
#  [9] "fraction.mt"               "Experiment"               
# [11] "Rank_nCount_RNA"           "HTO_maxID"                
# [13] "HTO_secondID"              "HTO_margin"               
# [15] "HTO_classification"        "HTO_classification.global"
# [17] "hash.ID"
```

```{r}
mergedSeurat
# An object of class Seurat 
# 20938 features across 15536 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: counts.Experiment1.1, counts.Experiment2.2
#  2 other assays present: HTO, CITE
```


### Demultiplex plots
Exploring the demultiplexing results with some plots

```{r demultiplex plots}

## Explore demultiplexing results
table(mergedSeurat$HTO_classification.global)
# some doublets and negatives were identified
# Doublet Negative  Singlet 
#     2055     1167    12314 

# ridgePlot
# check whether the demultiplex results agree with sample classification
RidgePlot(mergedSeurat,
          assay = "HTO",
          features = rownames(mergedSeurat[["HTO"]]),
          ncol = 2)

# heatmap

HTOHeatmap(mergedSeurat,
           assay = "HTO",
           ncells = 16000) #all cells <16k
```

```{r hto dimplots}
## Check out results in dimension reduction
mergedSeurat_HTO <- mergedSeurat

# subset to positive (singlet, doublet) HTO cells
mergedSeurat_HTO <- subset(mergedSeurat_HTO,
                           subset = HTO_classification.global != "Negative")

# change to HTO assay
DefaultAssay(mergedSeurat_HTO) <- "HTO"

# scale data
mergedSeurat_HTO <- ScaleData(mergedSeurat_HTO,
                              features = rownames(mergedSeurat_HTO),
                              verbose = FALSE)

# run PCA
mergedSeurat_HTO <- RunPCA(mergedSeurat_HTO, 
                           features = rownames(mergedSeurat_HTO),
                           approx = FALSE)

# run TSNE on PCs
mergedSeurat_HTO <- RunTSNE(mergedSeurat_HTO,
                            dims = 1:6, 
                            perplexity = 80,
                            check_duplicates=FALSE)

# plot 
## by singlet or droplets or negative 
DimPlot(mergedSeurat_HTO,
        group.by = "HTO_classification.global", 
        split.by="Experiment")

# by specific HTO identity (tissues here)
p <- DimPlot(mergedSeurat_HTO, 
        group.by = "HTO_classification", 
        split.by="Experiment",
        label = TRUE)

p + NoLegend()
```

  
There are some doublet clusters in both experiments, but the big groups are singlets nonetheless. We will filter for doublets in the next code chunks.

```{r, include=FALSE}
## Remove unneeded objects
rm(list=c("mergedSeurat_HTO", "Experiment1", "Experiment2", "xlim_upper"))
gc()
```

## CITE-seq antibody tags
Normalize the CITE-seq antibody-derived tags  in each experiment separately.

```{r CITEseq antibody tags}
## Subset for each individual experiment
Experiment1 <- subset(mergedSeurat, subset = Experiment == "Experiment1")
Experiment2 <- subset(mergedSeurat, subset = Experiment == "Experiment2")


## Normalize within each Experiment
Experiment1 <- NormalizeData(Experiment1, assay = "CITE", 
                             normalization.method = "CLR")

Experiment2 <- NormalizeData(Experiment2, assay = "CITE", 
                             normalization.method = "CLR")

## Merge the (biological) replicate experiments
mergedSeurat <- merge(x=Experiment1,
                      y=Experiment2) # don't add again cell id to avoid repetition

```

## Filter HTO doublets
Keep only singlets, filtering out the doublets and the negatives.


```{r}
## Filter for singlets
writeLines("Table before filtering:")
mergedSeurat$HTO_classification.global %>% table()
 # Doublet Negative  Singlet 
 #    2055     1167    12314 

# keep only singlet
mergedSeurat <- subset(mergedSeurat,
                       subset = HTO_classification.global == "Singlet")

writeLines("Table after filtering")
mergedSeurat$HTO_classification.global %>% table()
# Singlet 
#   12314 

```



***

# Cell Cycle

## Subset and normalize experiments
```{r}

## Subset for each individual experiment
Experiment1 <- subset(mergedSeurat, subset = Experiment == "Experiment1")
Experiment2 <- subset(mergedSeurat, subset = Experiment == "Experiment2")


## Normalize within each Experiment
Experiment1 <- NormalizeData(Experiment1, assay = "RNA", normalization.method = "LogNormalize")
Experiment2 <- NormalizeData(Experiment2, assay = "RNA", normalization.method = "LogNormalize")
```

## Cell cycle scores

### Human to mouse
```{r}
# Load Seurat’s built-in lists
s.genes <- Seurat::cc.genes.updated.2019$s.genes
g2m.genes <- Seurat::cc.genes.updated.2019$g2m.genes

# Convert to lowercase to match your data
s.genes <- tolower(s.genes)
g2m.genes <- tolower(g2m.genes)

# function to capitalize only first letter - mm 
capitalize_first <- function(x) {
  paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

s.genes <- capitalize_first(s.genes)
g2m.genes <- capitalize_first(g2m.genes)

g2m.genes
s.genes
```

### Calculate scores

```{r}
## Calculate cellcycle scores
Experiment1 <- CellCycleScoring(Experiment1,
                                 g2m.features = g2m.genes,
                                 s.features = s.genes)

Experiment2 <- CellCycleScoring(Experiment2,
                                 g2m.features = g2m.genes,
                                 s.features = s.genes)

## Merge the (biological) replicate experiments
# with the cell cycle info
mergedSeurat <- merge(x=Experiment1,
                      y=Experiment2)

mergedSeurat[["RNA"]] <- JoinLayers(mergedSeurat[["RNA"]])

# overview
mergedSeurat
# An object of class Seurat 
# 20938 features across 12314 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: data, counts
#  2 other assays present: HTO, CITE

# check metadata, new cols for cell cycle
# View(mergedSeurat@meta.data)
intersect(rownames(mergedSeurat), s.genes)
intersect(rownames(mergedSeurat), g2m.genes)

```


### Plot cell cycle markers
Should not be highly expressed

```{r}
Idents(mergedSeurat) <- mergedSeurat@meta.data$Phase

# Visualize the distribution of cell cycle markers across
RidgePlot(mergedSeurat, features = c("Pcna", "Top2a", "Mcm6", "Mki67"), ncol = 2)

RidgePlot(mergedSeurat, features = c("S.Score", "G2M.Score"), ncol = 1)

VlnPlot(mergedSeurat, features = c("S.Score", "G2M.Score"), pt.size = 0.1)

ggplot(mergedSeurat@meta.data, aes(x = S.Score, y = G2M.Score, color = Phase)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_bw()

```


```{r}
table(mergedSeurat$Phase)
cor(mergedSeurat$S.Score, mergedSeurat$G2M.Score)
# 0.4 it's ok

length(intersect(rownames(mergedSeurat), s.genes))
length(intersect(rownames(mergedSeurat), g2m.genes))

```

### PCA of cell cycle scores

Cell cycle phases are mixed and overlapping across the PCs, without clear separation of the phases. This shows that cell cycle is not driving cells expression.

```{r}
mergedSeurat2 <- NormalizeData(mergedSeurat, verbose = FALSE)

mergedSeurat2 <- FindVariableFeatures(mergedSeurat2, nfeatures = 3000, verbose = FALSE)

# scale for all features
mergedSeurat2 <- ScaleData(mergedSeurat2, verbose = FALSE)

# PCA on variable genes - to check whether cell cycle genes are influencing or not
mergedSeurat2 <- RunPCA(mergedSeurat2, npcs = 50, verbose = FALSE)

# plot
DimPlot(mergedSeurat2, reduction = "pca", group.by = "Phase", label = TRUE)

# clean
rm(mergedSeurat2)
gc()
```


# Save object
```{r}
## Save Seurat Object
filename <- file.path(dir_processed, "01_Intestinal_scRNAseq_postQC.rds")
filename

saveRDS(mergedSeurat, file = filename)
```

