---
title: "HDAC1 and HDAC2 Intestinal Immunity - scRNAseq T-cell identity and filtering"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean

---

# Cell identity of scRNAseq of mouse intestinal CD4+ and CD8+ T cells

Here we will go through an extra quality control step where we check for cell identity. We want to keep only relevant cells - T cells and NK cells. For that we will use SingleR prediction of cell types.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r libraries, message=FALSE,warning=FALSE}
library(here)
library(tidyverse)
library(patchwork)
library(Seurat)
library(SingleR)
library(celldex)
```

## Dir
```{r parameters}
dir_processed <- here("Data", "Processed", "01_QualityControl")
dir_output <- here("Data", "Processed", "02_TcellFiltering")

dirs <- c(dir_processed, dir_output)
lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```

## Load data
```{r}
input_file <- file.path(dir_processed, "01_Intestinal_scRNAseq_postQC.rds")
mergedSeurat <- LoadSeuratRds(input_file)

# overview
mergedSeurat
# An object of class Seurat 
# 20938 features across 12314 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: counts.Experiment1.1.1, counts.Experiment2.2.2
#  2 other assays present: HTO, CITE
```

## Cell type SingleR

Predict the cell types with SingleR reference and later filter the unexpected cell types.

```{r}
## Load reference dataset, see https://bioconductor.org/packages/release/bioc/vignettes/SingleR/inst/doc/SingleR.html
mouseRef <- celldex::MouseRNAseqData() 

## Subset for each individual experiment
Experiment1 <- subset(mergedSeurat, subset = Experiment == "Experiment1")
Experiment2 <- subset(mergedSeurat, subset = Experiment == "Experiment2")

## Predict cell type via singleR
pred.cells_1 <- SingleR(test = Experiment1@assays$RNA$data,
                        ref = mouseRef,
                        assay.type.test = "RNA",
                        assay.type.ref = "logcounts",
                        labels = mouseRef$label.main)

Experiment1$pred.cellType <- pred.cells_1$pruned.labels

pred.cells_2 <- SingleR(test = Experiment2@assays$RNA$data, 
                        ref = mouseRef, 
                        assay.type.test = "RNA",
                        assay.type.ref = "logcounts",
                        labels = mouseRef$label.main)

Experiment2$pred.cellType <- pred.cells_2$pruned.labels

## Merge the replicate experiments
mergedSeurat <- merge(x=Experiment1,
                      y=Experiment2)
# join layers
mergedSeurat[["RNA"]] <- JoinLayers(mergedSeurat[["RNA"]])

Idents(mergedSeurat) <- mergedSeurat@meta.data$pred.cellType

# overview
mergedSeurat
# An object of class Seurat 
# 20938 features across 12314 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: data, counts
#  2 other assays present: HTO, CITE

```

## Filter for only T cells and NK cells

```{r}
# check out cell types
table(mergedSeurat$pred.cellType)
# B cells Erythrocytes  Hepatocytes  Macrophages    Microglia    Monocytes     NK cells 
#           30            1            3            4            2            1         3256 
#      T cells 
#         9010 

mergedSeurat <- subset(mergedSeurat,
                       subset = pred.cellType == "T cells" | pred.cellType == "NK cells")
# check
table(mergedSeurat$pred.cellType)
# NK cells  T cells 
#     3256     9010 
mergedSeurat
# An object of class Seurat 
# 20938 features across 12266 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: data, counts
#  2 other assays present: HTO, CITE

## 48 irrelevant cells removed
```

## Save obj
```{r}
filename <- file.path(dir_output, "02_Intestinal_scRNAseq_Tcells.rds")
# filename

saveRDS(mergedSeurat, file = filename)
```

