---
title: "HDAC1 and HDAC2 Intestinal Immunity - scRNAseq Integration"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Biological replicates Integration

Here we will Integrate biological replicates using SCTransform-based integration, followed by dimensionality reduction and clustering. 
To check integrated data quality, we will perform Cell cycle scoring again, and Cell type predictions with SingleR Immgen reference as well.


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)

set.seed(42)
```


## Libraries
```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(Seurat)
library(patchwork)
library(celldex)
library(SingleR)

```

## Dir
```{r parameters}
dir_processed <- here("Data", "Processed", "02_TcellFiltering")
dir_integ <- here("Data", "Processed", "03_Integration")
dir_res <- here("Results", "03_Integration", "Plots")

dirs <- c(dir_processed, dir_integ, dir_res)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)

```

## Data
```{r}
input_file <- file.path(dir_processed, "02_Intestinal_scRNAseq_Tcells.rds")
# input_file

# load it
mergedSeurat <- LoadSeuratRds(input_file)

# overview
mergedSeurat
# An object of class Seurat 
# 20938 features across 12266 samples within 3 assays 
# Active assay: RNA (20929 features, 0 variable features)
#  2 layers present: data, counts
#  2 other assays present: HTO, CITE
# View(mergedSeurat@meta.data)

```

# SCTransform integration

## Subset experiments

```{r warning=FALSE}
IntegSeurat <- mergedSeurat

## Subset for each individual experiment and perform SC transform
Experiment1 <- subset(mergedSeurat, subset = Experiment == "Experiment1")
Experiment2 <- subset(mergedSeurat, subset = Experiment == "Experiment2")

Experiment1 <- SCTransform(Experiment1)
Experiment2 <- SCTransform(Experiment2)

Seurat_list <- list(Experiment1, Experiment2)

```

## Integrate features

```{r message=TRUE, warning=FALSE}
IntegFeatures <- SelectIntegrationFeatures(object.list = Seurat_list, nfeatures=3000)

Seurat_list <- PrepSCTIntegration(object.list = Seurat_list,
                                  anchor.features = IntegFeatures)

IntegAnchors <- FindIntegrationAnchors(object.list = Seurat_list, 
                                       normalization.method = "SCT",
                                       anchor.features = IntegFeatures)

IntegSeurat <- IntegrateData(anchorset = IntegAnchors,
                             normalization.method = "SCT")

```

# Check integrated obj

```{r}
IntegSeurat
# An object of class Seurat 
# 41166 features across 12266 samples within 5 assays 
# Active assay: integrated (3000 features, 3000 variable features)
#  2 layers present: data, scale.data
#  4 other assays present: RNA, HTO, CITE, SCT
```


## Plots before/after

```{r message=FALSE, warning=FALSE}
# pca before
IntegSeurat <- RunPCA(IntegSeurat, assay="SCT", features = VariableFeatures(IntegSeurat))
pca.bf <- DimPlot(IntegSeurat, reduction = "pca", group.by = c("Experiment")) + ggtitle("PCA before integration")

# umap before
IntegSeurat <- RunUMAP(IntegSeurat, dims=1:12)
umap.bf <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("Experiment")) + ggtitle("UMAP before Integration")

# pca after
IntegSeurat <- RunPCA(IntegSeurat, assay="integrated", features = VariableFeatures(IntegSeurat))
pca.af <- DimPlot(IntegSeurat, reduction = "pca", group.by = c("Experiment")) + ggtitle("PCA after integration")

# umap after
IntegSeurat <- RunUMAP(IntegSeurat, dims=1:12, min.dist = 0.2)
umap.af <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("Experiment")) + ggtitle("UMAP after Integration")

# clean
rm(list=c("mergedSeurat", "Seurat_list", "IntegAnchors", "IntegFeatures"))

# plots
(pca.bf | pca.af) / (umap.bf | umap.af)

```

## UMAP per tissue

```{r}
# add new col with tissue info from HTO classification
IntegSeurat$tissue <- IntegSeurat$HTO_classification %>%
  sub(x=., pattern = "WT-|KO-", replacement = "")

tissue <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("tissue")) + ggtitle("UMAP after Integration - HTO class")

tissue
```

## Save UMAP 
```{r}
ggsave(file.path(dir_res, "IntestinalSeurat_integrated_UMAP_tissue.png"),
       plot = tissue, device = "png", width = 10)
```


```{r}
## Plot with HTO overlay
check <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("HTO_classification")) + ggtitle("UMAP after Integration - HTO projection")

check

## Plot with singleR cell prediction overlay
check <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("pred.cellType")) + ggtitle("UMAP after Integration - singleR MouseRNAData prediction ")

check
```


## Elbowplot After  
```{r}
ElbowPlot(IntegSeurat, reduction="pca", ndims=50)

```
  
  
## Save  RDS
```{r}
filename <- file.path(dir_integ,"03-1_IntestinalSeurat_integrated.rds")
filename

saveRDS(IntegSeurat, file = filename)
```



***


# SingleR ImmGen

Check now in the integrated dataset how are the annotated cell types from the ImmGen ref data

```{r}
## Predict via singleR ImmGennData reference
# main labels
mouseRef <- celldex::ImmGenData()

IntegSeurat[["RNA"]] <- JoinLayers(IntegSeurat[["RNA"]])

pred.cells.main <- SingleR(test = IntegSeurat@assays$RNA$data,
                           ref = mouseRef,
                           assay.type.test = "RNA", 
                           assay.type.ref = "logcounts", 
                           labels = mouseRef$label.main)

IntegSeurat$pred.cellType_ImmGen_label.main <- pred.cells.main$labels

check <- DimPlot(IntegSeurat,
                 reduction = "umap", 
                 group.by = c("pred.cellType_ImmGen_label.main"),
                 label = TRUE) + 
  ggtitle("UMAP after Integration")

check

# how many cells from each group
table(IntegSeurat@meta.data$pred.cellType_ImmGen_label.main)
        #       DC Epithelial cells              ILC      Macrophages 
        #        1                1             4061                1 
        # NK cells              NKT          T cells              Tgd 
        #       26             1094             5292             1790 


## Predict via singleR ImmGennData reference
# fine labels
pred.cells.fine <- SingleR(test = IntegSeurat@assays$RNA$data,
                           ref = mouseRef,
                           assay.type.test = "RNA",
                           assay.type.ref = "logcounts",
                           labels = mouseRef$label.fine)

IntegSeurat$pred.cellType_ImmGen_label.fine <- pred.cells.fine$labels

check2 <- DimPlot(IntegSeurat, reduction = "umap",
                 group.by = c("pred.cellType_ImmGen_label.fine"),
                 label = FALSE) + ggtitle("UMAP after Integration") + NoLegend()

check2

```


```{r message=FALSE, warning=FALSE}
# clean
rm(list=c("mouseRef", "check", "pred.cells.main", "pred.cells.fine"))
gc()


```

## Save RDS

```{r}
filename <- file.path(dir_integ, "03-2_IntestinalSeurat_integrated_ImmGen.rds")
filename

saveRDS(IntegSeurat, file = filename)
```

  
# Cell cycle again  

Analyse considering the cell labels and in the integrated dataset

### Human to mouse
```{r}
# Load Seurat’s built-in lists
s.genes <- Seurat::cc.genes.updated.2019$s.genes
g2m.genes <- Seurat::cc.genes.updated.2019$g2m.genes

# Convert to lowercase to match your data
s.genes <- tolower(s.genes)
g2m.genes <- tolower(g2m.genes)

# function to capitalize only first letter - mm 
capitalize_first <- function(x) {
  paste0(toupper(substr(x, 1, 1)), substr(x, 2, nchar(x)))
}

s.genes <- capitalize_first(s.genes)
g2m.genes <- capitalize_first(g2m.genes)

g2m.genes
s.genes
```

### CellCycle scores

```{r}
## Calculate cellcycle scores
IntegSeurat <- CellCycleScoring(IntegSeurat,
                                 g2m.features = s.genes,
                                 s.features = g2m.genes)
IntegSeurat$S.Score %>% hist(main = "S Score")
IntegSeurat$G2M.Score %>% hist(main = "G2M Score")
IntegSeurat$Phase %>% table()

```

### Plots

#### Scores per phase

```{r}
Idents(IntegSeurat) <- IntegSeurat@meta.data$Phase

# Visualize the distribution of cell cycle markers across
RidgePlot(IntegSeurat, features = c("Pcna", "Top2a", "Mcm6", "Mki67"), ncol = 2)
# Warning: The following requested variables were not found: Pcna, Mcm6

RidgePlot(IntegSeurat, features = c("S.Score", "G2M.Score"), ncol = 1)

VlnPlot(IntegSeurat, features = c("S.Score", "G2M.Score"), pt.size = 0.1)
# very simmilar to previous but cleaner and less outlier cells 

ggplot(IntegSeurat@meta.data, aes(x = S.Score, y = G2M.Score, color = Phase)) +
  geom_point(alpha = 0.4, size = 0.6) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_bw()
# also here less diagonal cells

```
```{r}
table(IntegSeurat$Phase)
cor(IntegSeurat$S.Score, IntegSeurat$G2M.Score)
# 0.4 it's still the same, it's ok though

length(intersect(rownames(IntegSeurat), s.genes))
length(intersect(rownames(IntegSeurat), g2m.genes))
# less intersection with cell cycle genes
```

#### Umap/PCA
```{r}
## Visualize cell cycle effects
# pca integrated per phase
IntegSeurat <- RunPCA(IntegSeurat, 
                      assay="integrated",
                      features = VariableFeatures(IntegSeurat))
pca.cc <- DimPlot(IntegSeurat, reduction = "pca", group.by = c("Phase")) + ggtitle("PCA after Integration")

# umap integrated per phase
IntegSeurat <- RunUMAP(IntegSeurat,
                       dims=1:12,
                       min.dist = 0.2)
umap.cc <- DimPlot(IntegSeurat, reduction = "umap", group.by = c("Phase")) + ggtitle("UMAP after Integration")


# where are the high exp cc cells?
feat <- FeaturePlot(object = IntegSeurat, features = "S.Score")
FeaturePlot(object = IntegSeurat, features = "G2M.Score")

# looks like in a specific T cell pop
# Tgd (Gamma-delta T) cells
umap.imm <- DimPlot(IntegSeurat,
                 reduction = "umap", 
                 group.by = c("pred.cellType_ImmGen_label.main"),
                 label = TRUE) + 
  ggtitle("UMAP after Integration")


(pca.cc | feat ) / (umap.cc | umap.imm )

umap.imm

```


## CITE-seq

Check protein abundance from CITEseq prots in the integrated obj 

```{r}
## Change default assay to cite
DefaultAssay(IntegSeurat) <- "CITE"


## Plot feature plot for each CITE-seq protein
for (i in rownames(IntegSeurat@assays$CITE$data)){
  print(i)
  p <- FeaturePlot(IntegSeurat, i, cols=c("lightgrey", "darkblue"), slot="data") + ggtitle(paste0("CITE-seq abundance: ", i))
  print(p)
}


## Change default assay to "cite"integrated" again
DefaultAssay(IntegSeurat) <- "integrated"
```


# Clustering

```{r}
## Calculate PCA again and perform clustering
IntegSeurat <- RunPCA(IntegSeurat,
                      assay="integrated",
                      features = VariableFeatures(IntegSeurat))

DimPlot(IntegSeurat,
        reduction = "pca",
        group.by = c("Experiment")) + ggtitle("PCA afterintegration")

IntegSeurat <- FindNeighbors(IntegSeurat, dims = 1:12)

IntegSeurat <- FindClusters(IntegSeurat, resolution = 0.5, verbose = FALSE)

## Visualize Clustering Resuls
DimPlot(IntegSeurat,
        reduction = "umap",
        group.by = c("seurat_clusters")) + ggtitle("UMAP: Clustering with res=0.5")
```



## Save 

```{r}
filename <- file.path(dir_integ, "03-3_IntestinalSeurat_integrated_clustered.rds")
filename

saveRDS(IntegSeurat, file = filename)
```

