---
title: "HDAC1 and HDAC2 Intestinal Immunity - scRNAseq Overview"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Overview of integrated object 

Descriptive and visual overview of previously integrated object, to generate the figures necessary for the manuscript.


# Libraries
```{r, warning = FALSE, message=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(Seurat)
library(biomaRt)
```

# Object source

By default, manuscript figures are reproduced using the frozen Seurat object used for the final analysis. Set `use_frozen_object = FALSE` to use the object generated by the preceding pipeline scripts.

*Minor differences in stochastic dimensionality-reduction results may occur when the full workflow is rerun from raw data.*

```{r}
use_frozen_object <- TRUE
```

## Dirs
```{r}
dir_biomart <- here("Data", "Processed", "Additional_metadata")
dir_input <- here("Data", "Processed", "03_Integration")
dir_frozen <- here("Data", "Processed", "00_Frozen")
dir_res <- here("Results", "04_Overview", "Plots")

dirs <- c(dir_res, dir_biomart)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```

# Data
```{r}
frozen_obj_file <- file.path(dir_frozen, "03-3_IntestinalSeurat_integrated_clustered_FROZEN.rds")

obj_file <- file.path(dir_input, "03-3_IntestinalSeurat_integrated_clustered.rds")

if (use_frozen_object) {
  obj <- LoadSeuratRds(frozen_obj_file)
  message("Using frozen Seurat obj for exact manuscript figure reproduction.")
} else {
  obj <- LoadSeuratRds(obj_file)
  message("Using Seurat object generated by the reproducible analysis pipeline.")
}
```

```{r}
obj
# An object of class Seurat 
# 41166 features across 12266 samples within 5 assays 
# Active assay: integrated (3000 features, 3000 variable features)
#  2 layers present: data, scale.data
#  4 other assays present: RNA, HTO, CITE, SCT
#  2 dimensional reductions calculated: pca, umap
```

## Tidy for plots 

Tidy frozen object for downstream plots 

```{r}

if (use_frozen_object) {
  # change order of HTO classes, as factor with specific levels
  obj$HTO_classification <- factor(obj$HTO_classification,
                                   levels = c("WT-colon", "KO-colon",
                                              "WT-SI-IEL", "KO-SI-IEL",
                                              "WT-SI-LPL", "KO-SI-LPL"))
  # add new col with tissue info from HTO classification
  obj$tissue <- obj$HTO_classification %>% sub(pattern = "WT-|KO-", replacement = "")
  # and genotype
  obj$genotype <- obj$HTO_classification %>% substr(1, 2) # takes the first two characters

  obj$tissue <- as.factor(obj$tissue)
  obj$genotype <- as.factor(obj$genotype)

  # names
  # collapse Exp1 repeats of the first underscore-delimited token
  fix_prefix <- function(x) sub("^(([^_]+))(?:_\\1)+_", "\\1_", x, perl = TRUE)
  # apply to cell names in obj
  new_names <- fix_prefix(colnames(obj))
  # check
  # head(new_names)
  
  # apply to obj and keep everything in sync
  colnames(obj) <- new_names
  rownames(obj@meta.data) <- new_names
  }

# View(obj@meta.data)
# colnames(obj@meta.data)
```


## Get biotype info from ensembl

```{r}
ensembl <- useEnsembl(biomart="genes", dataset = "mmusculus_gene_ensembl")

ensembl
# Object of class 'Mart':
#   Using the ENSEMBL_MART_ENSEMBL BioMart database
#   Using the mmusculus_gene_ensembl dataset

# which ensembl version used
listEnsembl()
# biomart                version
# 1         genes      Ensembl Genes 115
# 2 mouse_strains      Mouse strains 115
# 3          snps  Ensembl Variation 115
# 4    regulation Ensembl Regulation 115
# release 115

biomart_tab <- getBM(attributes = c("external_gene_name",
                                    "description", "gene_biotype"),
                     filters = "external_gene_name",
                     values = rownames(obj@assays$RNA),
                     mart = ensembl)

```

## Save biomart
```{r}
write.csv(biomart_tab, file = file.path(dir_biomart, "Ensembl_115_biomart_mm10_biotype.csv"))
```


***


# Overview of all cells


## Descriptive overview

```{r}
# Print Experiments
writeLines("Experiments:")
obj$Experiment %>% unique %>% sort()
# [1] "Experiment1" "Experiment2"


writeLines("Cells per experiment:")
obj$Experiment %>% table() %>% print()
# Experiment1 Experiment2 
#        3802        8464 

## Print Tissues
writeLines("Tissues:")
obj@meta.data$tissue %>% unique() %>% as.character() %>% sort()
# [1] "colon"  "SI-IEL" "SI-LPL"

## Genotypes
writeLines("Genotype:")
obj@meta.data$genotype %>% unique() %>% as.character() %>% sort()
# [1] "KO" "WT"

## Print HTO classes
writeLines("HTO classes:")
obj@meta.data$HTO_classification %>% unique() %>% as.character() %>% sort()
# [1] "KO-colon"  "KO-SI-IEL" "KO-SI-LPL" "WT-colon"  "WT-SI-IEL" "WT-SI-LPL"

writeLines("Cells per HTO class:")
obj@meta.data$HTO_classification %>% table() %>% print()
 # WT-colon  KO-colon WT-SI-IEL KO-SI-IEL WT-SI-LPL KO-SI-LPL 
 #     1266      1568      3596      2475      1352      2009 

## Print CITE-seq quantified proteins
writeLines("CITE-seq targets:")
obj@assays$CITE$counts %>% rownames()
# [1] "CD4"  "CD8a" "CD8b"
```


## Plots overview


```{r}

## Show Dimension reduction by Experiment
DimPlot(obj, reduction = "umap", group.by = c("Experiment")) + ggtitle("UMAP colored by Experiment")

DimPlot(obj, reduction = "umap", group.by = c("Experiment"), split.by = c("Experiment")) + ggtitle("UMAP split by Experiment")

## Show Dimension reduction by Tissue
p1 <-DimPlot(obj, reduction = "umap", group.by = c("tissue")) + ggtitle("UMAP colored by tissue origin")
print(p1)

# save
p1file <- file.path(dir_res,"IntestinalSeurat_integrated_UMAP_tissue.png")
p1file

ggsave(p1file, plot = p1, device = "png", width = 10)


# Show Dimension reduction by HTO_class
p2 <- DimPlot(obj, reduction = "umap", group.by = c("HTO_classification")) + ggtitle("UMAP colored by HTO class")
print(p2)

# save
p2file <- file.path(dir_res,"IntestinalSeurat_integrated_UMAP_HTO_class.png")
p2file
ggsave(p2file, plot = p2, device = "png", width = 10, height = 6)


```


## SingleR annotations
```{r}
## Whole obj
p3 <-DimPlot(obj, reduction = "umap", group.by = c("pred.cellType")) + ggtitle("UMAP colored by predicted Cell Type (SingleR/celldex)")
print(p3)

# save
p3file <- file.path(dir_res, "IntestinalSeurat_integrated_UMAP_CellType.png")
p3file
ggsave(p3file, plot = p3, device = "png")


## Split by HTO class
p4 <-DimPlot(obj, reduction = "umap", group.by = c("pred.cellType"), split.by = "HTO_classification") + ggtitle("UMAP grouped by predicted Cell Type and split by HTO class")
print(p4)

p4file <- file.path(dir_res,"IntestinalSeurat_integrated_UMAP_CellType_HTOSplit.png")
p4file
ggsave(p4file, plot = p4, device = "png", width = 30)

```

## Integrated Clusters

```{r}
# just cheking
identical(obj@meta.data$integrated_snn_res.0.5, obj@meta.data$seurat_clusters)

clust <-DimPlot(obj, reduction = "umap", group.by = "seurat_clusters") + ggtitle("UMAP by integrated clusters")
print(clust)

# save
clustf <- file.path(dir_res,"IntestinalSeurat_integrated_UMAP_Clusters.png")
clustf
ggsave(clustf, plot = clust, device = "png", width = 10)


### with labels
clust2 <-DimPlot(obj, reduction = "umap", group.by = "seurat_clusters", label = TRUE) + ggtitle("UMAP labeled by integrated clusters ")
print(clust2)

# save
clust2f <- file.path(dir_res,"IntestinalSeurat_integrated_UMAP_ClustersLabels.png")
clust2f
ggsave(clust2f, plot = clust2, device = "png", width = 10)


```


## CITEseq 

### Individual plots

```{r}
## Change default assay to CITE
DefaultAssay(obj) <- "CITE"

## Get all CITE features from the data slot
cite_features <- rownames(obj@assays$CITE$data)

## Plot + save feature plot for each CITE-seq protein
for (i in cite_features) {
  message("Plotting: ", i)
  
  p <- FeaturePlot(
    obj, 
    features = i,
    cols = c("lightgrey", "darkblue"),
    slot = "data"
  ) + ggtitle(paste0("CITE-seq abundance: ", i))
  
  print(p)
  
  # File name following convention: CITE-seq abundance + feature name
  # IntestinalSeurat_integrated_FeaturePlot_CITEseq_<protein>_<date>.png
  cite_file <- file.path(
    dir_res,
    paste0(
      "IntestinalSeurat_integrated_FeaturePlot_CITEseq_",
      i, ".png")
  )
  
  ggsave(
    filename = cite_file,
    plot     = p,
    device   = "png",
    width    = 10
  )
}



```

### Multipanel
```{r}
## Vector of the 3 proteins 
proteins_cite <- rownames(obj@assays$CITE$data) 

## One multipanel FeaturePlot
p_cite_multi <- FeaturePlot(
  obj,
  features = proteins_cite,
  cols     = c("lightgrey", "darkblue"),
  slot     = "data",
  ncol = 3
)

print(p_cite_multi)

## Save 
multi_file <- file.path(
  dir_res,
  "IntestinalSeurat_integrated_FeaturePlot_CITEseq_AllProts.png"
)

ggsave(
  filename = multi_file,
  plot     = p_cite_multi,
  device   = "png",
  width    = 30
)

```
