---
title: "HDAC1 and HDAC2 Intestinal Immunity - scRNAseq Subset HTO classes"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Analysis of HTO classes: subsets of intestinal tissues

There are three different tissues to be investigated:

1. Small intestinal intraepithelial **(SI-IEL)**

2. Small intestinal lamina propria **(SI-LPL)**

3. Colon


In this script it is possible to select one out of the three tissues and run all donwstream analysis. See the *parameters* section.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries  

```{r, message=FALSE, warnings=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(knitr)
library(Seurat)
```

# Object source

By default, manuscript figures are reproduced using the frozen Seurat object used for the final analysis. Set `use_frozen_object = FALSE` to use the object generated by the preceding pipeline scripts.

*Minor differences in stochastic dimensionality-reduction results may occur when the full workflow is rerun from raw data.*

```{r}
use_frozen_object <- TRUE

```

## Directories
Set up relevant dirs for this script

```{r}
dir_input <- here("Data", "Processed", "03_Integration")
dir_frozen <- here("Data", "Processed", "00_Frozen")

frozen_obj_file <- file.path(dir_frozen, "03-3_IntestinalSeurat_integrated_clustered_FROZEN.rds")
obj_file <- file.path(dir_input, "03-3_IntestinalSeurat_integrated_clustered.rds")

if (use_frozen_object) {
  obj <- LoadSeuratRds(frozen_obj_file)
  message("Using frozen Seurat obj for exact manuscript figure reproduction.")
  
} else {
  obj <- LoadSeuratRds(obj_file)
  message("Using Seurat object generated by the reproducible analysis pipeline.")
  
}

# inspect obj
obj
# An object of class Seurat 
# 41166 features across 12266 samples within 5 assays 
# Active assay: integrated (3000 features, 3000 variable features)
#  2 layers present: data, scale.data
#  4 other assays present: RNA, HTO, CITE, SCT
#  2 dimensional reductions calculated: pca, umap
```

## Tidy for plots

Tidy frozen obj for downstream plots 

```{r}
if (use_frozen_object) {
  obj$HTO_classification <- factor(obj$HTO_classification,
                                   levels = c("WT-colon", "KO-colon",
                                              "WT-SI-IEL", "KO-SI-IEL",
                                              "WT-SI-LPL", "KO-SI-LPL"))
  
  obj$tissue <- obj$HTO_classification %>% sub(pattern = "WT-|KO-", replacement = "")
  obj$genotype <- obj$HTO_classification %>% substr(1, 2)

  obj$tissue <- as.factor(obj$tissue)
  obj$genotype <- as.factor(obj$genotype)
}
```



## Parameters
Specify which HTO classes should be analyzed further separately.

### Select tissue [edit here!]

```{r}
## Specify which HTO class to be analyzed separately.
## Option 1:
TISSUE <- "SI-IEL" 

## Option 2:
# TISSUE <- "SI-LPL" 

## Option 3:
# TISSUE <- "colon" 

stopifnot(TISSUE %in% c("SI-IEL", "SI-LPL", "colon"))
 
message("(!) Attention! The chosen tissue for this script is: ", TISSUE)

```

```{r}
# map the tissue to the specific HTO classes considering the genotype
tissue_dict <- list(
  "SI-IEL" = c("WT-SI-IEL",  "KO-SI-IEL"),
  "SI-LPL" = c("WT-SI-LPL",  "KO-SI-LPL"),
  "colon"  = c("WT-colon",   "KO-colon")
)
tissue_dict

```

### Results directories
Will change depending on the tissue

```{r}
OUTDIR <- here("Results", "05_Subset", TISSUE, "Plots")
SUBDIR_OBJ <- here("Data", "Processed", "05_Subset", TISSUE)

dirs <- c(OUTDIR, SUBDIR_OBJ)
# create dir if it does not exist yet
lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)


```

### Path helper
```{r}
## helper for making paths in this dir structure
make_path <- function(dir, filename) file.path(dir, paste0(TISSUE,"_", filename))

# usage like
# saveRDS(obj_sub, make_path(SUBDIR_OBJ, "seurat_subset.rds"))
# ggsave(make_path("umap_clusters.pdf"), p_umap)
# write.csv(obj_sub@meta.data, make_path("metadata.csv"))
```


### Subset parameters
```{r}
# get the specific WT/KO from the selected tissue
subselection_HTOclasses <- tissue_dict[[TISSUE]]
# check
message("These are the genotypes measured in this tissue: ")
print(subselection_HTOclasses)

# generate groups for Differential Analysis pairwise comparisons in correct order (KO first, WT second)
DEgroups_HTOclasses <- list(
  list(
    c(subselection_HTOclasses[2]),   # KO first
    c(subselection_HTOclasses[1])    # WT second
  )
)
message("These are the pairwise comparison for further DE analysis:")
print(DEgroups_HTOclasses)
```

### UMAP/Cluster parameters
```{r}
# randomness fix
set.seed(42)


## Specify the number of PCA dimensions to be used for 1) dimension reduction via UMAP, and 2) clustering
numberDimsPCA_for_UMAP = 10

## Specify the minimum distance (first main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
minDist_for_UMAP = 0.1

## Specify the spread (second main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
spread_for_UMAP = 1

## Specify number of nearest neighbours for dimension reduction via UMAP
n_neighbours_for_UMAP = 30

## Specify local connectivity for dimension reduction via UMAP
local_connectivity_for_UMAP = 20

## Specify negative sample rate for dimension reduction via UMAP
negative_sample_rate_for_UMAP = 20

## Specify the resolution to be used for clustering. A lower resolution will produce a lower number of clusters
res_for_clustering = 0.5

```



***


# Subset obj
```{r}
## Perform subselection
obj.tissue <- subset(obj,
                     subset = HTO_classification %in% subselection_HTOclasses)

obj.tissue
# An object of class Seurat 
# 41166 features across 6071 samples within 5 assays 
# Active assay: integrated (3000 features, 3000 variable features)
#  2 layers present: data, scale.data
#  4 other assays present: RNA, HTO, CITE, SCT
#  2 dimensional reductions calculated: pca, umap
```

```{r}
## Redefine HTO classification as factor
obj.tissue$HTO_classification <- factor(obj.tissue$HTO_classification,
                                        levels = subselection_HTOclasses)

```


## Overview
```{r}
# Print Experiments
writeLines("Experiments:")
obj.tissue$Experiment %>% unique %>% sort()
# [1] "Experiment1" "Experiment2"

writeLines("Cells per experiment:")
obj.tissue$Experiment %>% table() %>% print()
# Experiment1 Experiment2 
#        1885        4186 

## Print Tissues
writeLines("Tissues:")
obj.tissue@meta.data$tissue %>% unique() %>% as.character() %>% sort()

## Genotypes
writeLines("Genotype:")
obj.tissue@meta.data$genotype %>% unique() %>% as.character() %>% sort()
# [1] "KO" "WT"

## Print HTO classes
writeLines("HTO classes:")
obj.tissue@meta.data$HTO_classification %>% unique() %>% as.character() %>% sort()
# [1] "KO-SI-IEL" "WT-SI-IEL"

writeLines("Cells per HTO class:")
obj.tissue@meta.data$HTO_classification %>% table() %>% print()
# WT-SI-IEL KO-SI-IEL 
     # 3596      2475 
     
## Print CITE-seq quantified proteins
writeLines("CITE-seq targets:")
obj.tissue@assays$CITE$counts %>% rownames()
# [1] "CD4"  "CD8a" "CD8b"
```

***

# Reintegration of subset

Note that the integration step of the two experiments needs to be performed again as other genes might explain most of the variation in the new subset of cells. In other words, the set of most-variable genes might differ from the original set of genes, on which basis PCA and UMAP are calculated. Here are the results after the new integration and subsequent recalculation of UMAP dimension reduction and PCA:

```{r}
## Delete previous "SCT" and "integrated" assays
DefaultAssay(obj.tissue) <- "RNA"

obj.tissue[["SCT"]] <- NULL
obj.tissue[["integrated"]] <- NULL


## Subset for each individual experiment and perform SC transform
Experiment1 <- subset(obj.tissue, subset = Experiment == "Experiment1")
Experiment2 <- subset(obj.tissue, subset = Experiment == "Experiment2")

Experiment1 <- SCTransform(Experiment1)
Experiment2 <- SCTransform(Experiment2)

obj.tissue_list <- list(Experiment1, Experiment2)

```

```{r message=FALSE, warning=FALSE, include = FALSE}
# clean
rm(list=c("Experiment1", "Experiment2"))
gc()
```

## SCT integration

```{r message=FALSE, warning=FALSE}
##  Perform Integration
IntegFeatures <- SelectIntegrationFeatures(object.list = obj.tissue_list,
                                           nfeatures=3000)

obj.tissue_list <- PrepSCTIntegration(object.list = obj.tissue_list, 
                                      anchor.features = IntegFeatures)

IntegAnchors <- FindIntegrationAnchors(object.list = obj.tissue_list,
                                       normalization.method = "SCT",
                                       anchor.features = IntegFeatures)
## new integrated seurat obj
int.obj.tissue <- IntegrateData(anchorset = IntegAnchors,
                                 normalization.method = "SCT")
```

### Check

```{r}
# after integration
DefaultAssay(int.obj.tissue)
# [1] "integrated"

Assays(int.obj.tissue)
# [1] "RNA"        "HTO"        "CITE"       "SCT"        "integrated"

nrow(GetAssayData(int.obj.tissue, assay = "SCT"))
# [1] 15709

nrow(GetAssayData(int.obj.tissue, assay = "integrated"))
# 3000

length(IntegFeatures)
# 3000
all(IntegFeatures %in% rownames(int.obj.tissue[["integrated"]]))
# TRUE

head(VariableFeatures(int.obj.tissue[["SCT"]]))
# 0
head(VariableFeatures(int.obj.tissue[["integrated"]]))
# features

```

### Compute missing SCT variableFEatures

We need SCT variable features to be able to compute before and after PCAs and UMAPs correctly.

Because this is a V4 integrated approach ina V5 type of object and Seurat version, we can't compute variable features with the normal FindVariableFEatures. This object here is multi-model - one SCT model per experiment (we have 2 exp).

So we will copy the VarFeat from the integrated layer: *“For the SCT assay in this integrated object, mark as variable features those genes that were used for integration (IntegFeatures) and that actually exist in this assay.”*

```{r}
# assign the integration features as variable features of the SCT assay
VariableFeatures(int.obj.tissue[["SCT"]]) <- IntegFeatures

head(VariableFeatures(int.obj.tissue[["SCT"]]))
# [1] "Gzma"    "Ccl5"    "Gm42418" "Gzmb"    "Xcl1"    "Frmd5"  
```


# Plots overview

## Before integration 
assay is SCT

```{r}
## PCA/UMAP before & after integration for one tissue

# PCA before (assay = "SCT")
DefaultAssay(int.obj.tissue) <- "SCT"

# RunPCA
int.obj.tissue <- RunPCA(
  int.obj.tissue,
  assay   = "SCT",
  features = VariableFeatures(int.obj.tissue[["SCT"]]),
  reduction.name = "pca.unintegrated"   # save it to obj
)

# DimPlot
pca.bf <- DimPlot(
  int.obj.tissue,
  reduction = "pca.unintegrated",
  group.by  = "Experiment"
) + ggtitle(paste("PCA before integration -", TISSUE))

ggsave(
  filename = file.path(OUTDIR, paste0(TISSUE, "_PCA_before_integration.png")),
  plot = pca.bf, width = 10)

# UMAP before
# RunUmap
int.obj.tissue <- RunUMAP(
  int.obj.tissue,
  reduction            = "pca.unintegrated", # which one to use
  reduction.name       = "umap.unintegrated", # which name to store it
  dims                 = 1:numberDimsPCA_for_UMAP, # parameter set up in Parameters
  min.dist             = minDist_for_UMAP, # also set
  spread               = spread_for_UMAP, # set before
  n.neighbors          = n_neighbours_for_UMAP, # set 
  local.connectivity   = local_connectivity_for_UMAP, # set
  negative.sample.rate = negative_sample_rate_for_UMAP # set
)


# DimPlot
umap.bf <- DimPlot(
  int.obj.tissue,
  reduction = "umap.unintegrated",
  group.by  = "Experiment"
) + ggtitle(paste("UMAP before integration -", TISSUE))

ggsave(
  filename = file.path(OUTDIR, paste0(TISSUE, "_UMAP_before_integration.png")),
  plot = umap.bf, width = 10)
```

### After integration
assay is integrated
```{r}
####### AFTER integration
# PCA after (assay = "integrated")
DefaultAssay(int.obj.tissue) <- "integrated"

# RunPCA
int.obj.tissue <- RunPCA(
  int.obj.tissue,
  assay    = "integrated",
  reduction.name = "pca.integrated",   # save it to obj
  features = VariableFeatures(int.obj.tissue[["integrated"]])
)

pca.af <- DimPlot(
  int.obj.tissue,
  reduction = "pca.integrated",
  group.by  = "Experiment"
) + ggtitle(paste("PCA after integration -", TISSUE))

ggsave(
  filename = file.path(OUTDIR, paste0(TISSUE, "_PCA_after_integration.png")),
  plot = pca.af, width = 10)

# UMAP AFTER
# Run UMAP
int.obj.tissue <- RunUMAP(
  int.obj.tissue,
  reduction            = "pca.integrated", # which one to use
  reduction.name       = "umap.integrated", # which name to save
  dims                 = 1:numberDimsPCA_for_UMAP,
  min.dist             = minDist_for_UMAP,
  spread               = spread_for_UMAP,
  n.neighbors          = n_neighbours_for_UMAP,
  local.connectivity   = local_connectivity_for_UMAP,
  negative.sample.rate = negative_sample_rate_for_UMAP
)

# DImplot
umap.af <- DimPlot(
  int.obj.tissue,
  reduction = "umap.integrated",
  group.by  = "Experiment"
) + ggtitle(paste("UMAP after integration -", TISSUE))

ggsave(
  filename = file.path(OUTDIR, paste0(TISSUE, "_UMAP_after_integration.png")),
  plot = umap.af, width = 10)

```

### All plots
```{r}
overview.pcas <- pca.bf | pca.af

ggsave(filename = file.path(OUTDIR, 
                            paste0(TISSUE, "_PCAs_before_after_overview.png")),
       plot = overview.pcas, width = 20)


overview.umaps <- umap.bf | umap.af

ggsave(filename = file.path(OUTDIR, paste0(TISSUE, "_UMAPs_before_after_overview.png")),
       plot = overview.umaps,  width = 20)

pca.bf
pca.af

umap.bf
umap.af


```


### ElbowPlot
```{r}
writeLines(paste("Elbow plot to determine appropriate number of PCA dimensions used for UMAP (", numberDimsPCA_for_UMAP, ")"))


ElbowPlot(int.obj.tissue, reduction="pca.integrated", ndims=50) + ggtitle(paste("Elbow plot integrated", TISSUE))

```

### Check obj
```{r}
int.obj.tissue
```

### JoinLayers
```{r}
int.obj.tissue[["RNA"]] <- JoinLayers(int.obj.tissue[["RNA"]])
```


### Save RDS
```{r}
myfile <- file.path(SUBDIR_OBJ, paste0("05_", TISSUE, "_IntestinalSeurat_TissueSubsetReintegrated.rds"))
myfile

saveRDS(int.obj.tissue, file = myfile)
```



## CITE-seq proteins

Note: CITE-seq protein abundance is visualized on the RNA-based integrated UMAP.

```{r}
DefaultAssay(int.obj.tissue) <- "CITE"

for (i in rownames(int.obj.tissue@assays$CITE$data)) {
  message("Processing: ", i)

  # Generate feature plot
  p <- FeaturePlot(
    int.obj.tissue,
    features = i,
    cols = c("lightgrey", "darkblue"),
    slot = "data"
  ) + ggtitle(paste0("", i))

  # Show plot in session
  print(p)

  # Save each plot
  ggsave(
    filename = make_path(dir = OUTDIR, filename = paste0("FeaturePlot_CITEseq_", i, ".png")),
    plot = p,
    width = 8, height = 6
  )

  # Clean up
  rm(p)
  gc()
}


```
