---
title: "HDAC1 and HDAC2 Intestinal Immunity - Subsets Clustering"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran & Monika Waldherr"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Gene investigation of subsets

Here we do:
- calculate the module scores based on a custom gene set for the intestinal tissue subset
- cluster the groups 

The intestinal tissue subsets can be three different types:

1. Small intestinal intraepithelial **(SI-IEL)**

2. Small intestinal lamina propria **(SI-LPL)**

3. Colon

The previous script `05_Intestinal_scRNAseq_subset.Rmd` produces the RDS object that is necessary for this downstream analysis.

Here it is possible to select one out of the three tissues and run all donwstream analysis. See the *parameters* section.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries  
```{r, message=FALSE, warnings=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(RColorBrewer)
library(knitr)
library(Seurat)
library(SeuratData)

library(ggalluvial)
library(grid)

```

# TISSUE parameter
```{r}
TISSUE <- "SI-IEL"

stopifnot(TISSUE %in% c("SI-IEL", "SI-LPL", "colon"))

```

# Object source

By default, manuscript figures are reproduced using the frozen Seurat object used for the final analysis. Set `use_frozen_object = FALSE` to use the object generated by the preceding pipeline scripts.

*Minor differences in stochastic dimensionality-reduction results may occur when the full workflow is rerun from raw data.*

```{r}
use_frozen_object <- TRUE
```


# Directories
```{r}
dir_input <- here("Data", "Processed", "05_Subset", TISSUE)
dir_frozen <- here("Data", "Processed", "00_Frozen")

dir_res <- here("Results", "06_ModuleScores_MarkerGenes", TISSUE)
dir_obj <- here("Data", "Processed", "06_ModuleScores", TISSUE)
dir_plots <- file.path(dir_res, "Plots")
dir_tables <- file.path(dir_res, "Tables")

dirs <- c(dir_obj, dir_plots, dir_tables)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```
## Read obj
```{r}
frozen_obj_file <- file.path(dir_frozen, "05_SI-IEL_IntestinalSeurat_TissueSubsetReintegrated_FROZEN.rds")

obj_file <- file.path(dir_input, paste0("05_", TISSUE, "_IntestinalSeurat_TissueSubsetReintegrated.rds"))

if (use_frozen_object) {
  stopifnot(TISSUE == "SI-IEL")
  obj <- LoadSeuratRds(frozen_obj_file)
  message("Using frozen Seurat obj for exact manuscript figure reproduction.")
} else {
  obj <- LoadSeuratRds(obj_file)
  message("Using Seurat object generated by the reproducible analysis pipeline.")
}

```

```{r}
obj

```

## Helper Functions
Helper function for creating new paths for new output

```{r}
## Make path
make_path <- function(dir, filename) file.path(dir, paste0(TISSUE,"_", filename))

# save all extensions 
save_plot <- function(plot, dir, name, w = 10, h = 6) {
  base <- file.path(dir, paste0(TISSUE, "_", name))
  ggsave(paste0(base, ".pdf"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".png"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".svg"), plot = plot, width = w, height = h)
}

```

## UMAP/Cluster parameters
```{r}
# randomness fix
set.seed(42)

## Specify the number of PCA dimensions to be used for 1) dimension reduction via UMAP, and 2) clustering
numberDimsPCA_for_UMAP = 10

## Specify the minimum distance (first main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
minDist_for_UMAP = 0.1

## Specify the spread (second main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
spread_for_UMAP = 1

## Specify number of nearest neighbours for dimension reduction via UMAP
n_neighbours_for_UMAP = 30

## Specify local connectivity for dimension reduction via UMAP
local_connectivity_for_UMAP = 20

## Specify negative sample rate for dimension reduction via UMAP
negative_sample_rate_for_UMAP = 20

## Specify the resolution to be used for clustering. A lower resolution will produce a lower number of clusters
res_for_clustering = 0.5

```


## Genes for Module score
```{r}
# Module Scores genes for annotation
# Load your CSV file (with cluster_name in column 1 and gene_name in column 2)
fileannot <- here("Data", "Processed", "Additional_metadata", "Annotation_final_list_for_module_score.csv")

geneSet_df <- read.csv(fileannot, header = FALSE, stringsAsFactors = FALSE)

```


***

## 3D CITE-seq

```{r}
DefaultAssay(obj) <- "CITE"

for (i in rownames(obj@assays$CITE$data)) {
  message("Processing: ", i)

  # Generate feature plot
  p <- FeaturePlot(
    obj,
    features = i,
    cols = c("lightgrey", "darkblue"),
    slot = "data"
  ) + ggtitle(paste0("", i))

  # Show plot in session
  print(p)

  # Save each plot
  ggsave(
    filename = make_path(dir = dir_plots, filename = paste0("FeaturePlot_CITEseq_", i, ".png")),
    plot = p,
    width = 8, height = 6
  )

  # Clean up
  rm(p)
  gc()
}


```


***



## Module Scores

Create Module Scores based on Custom Gene Sets for annotaion


### JoinLayers
```{r}
# make sure layers are joined
obj[["RNA"]] <- JoinLayers(obj[["RNA"]])

```


```{r}
# Set default assay
DefaultAssay(obj) <- "RNA"

## Extract feature names
featurenames <- obj@assays$RNA$data %>% rownames()
# featurenames_integrated <- obj@assays$integrated$data %>% rownames()


# Optional: Clean whitespace
geneSet_df$cluster_name <- trimws(geneSet_df[[1]])
geneSet_df$gene_name <- trimws(geneSet_df[[2]])

# Remove empty rows or missing genes
geneSet_df <- geneSet_df[geneSet_df$gene_name != "" & !is.na(geneSet_df$gene_name), ]

# Group genes by cluster
gene_sets <- split(geneSet_df$gene_name, geneSet_df$cluster_name)

# Loop through each gene set (per cluster)
for (cluster in names(gene_sets)) {
  
  customGeneSet <- gene_sets[[cluster]]
  
  # Filter genes to those present in the dataset
  customGeneSet <- customGeneSet[customGeneSet %in% featurenames]
  
  if (length(customGeneSet) == 0) {
    message(paste0("No matching genes found in dataset for cluster: ", cluster))
    next
  }
  
  # Print info
  message("Cluster: ", cluster)
  message("Number of genes found in data: ", length(customGeneSet))
  print(head(customGeneSet, 100))
  
  # Calculate module score
  obj <- AddModuleScore(
    object = obj,
    features = list(customGeneSet),
    name = cluster,
    assay = "RNA",
    slot = "data"
  )
  
  # Create FeaturePlot
  feature_to_plot <- paste0(cluster, "1")
  
  gg <- FeaturePlot(
    object = obj,
    features = feature_to_plot
  ) + 
    scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
    ggtitle(cluster) +
    theme(
      plot.title = element_text(size = 35,
                                face = "plain",
                                hjust = 0.5,
                                vjust = -35), # centered and placed below
    legend.position = "none" , # removes the legend
    axis.title = element_blank()
  )
  
 
  # Save plot
  filename <- make_path(dir = dir_plots, filename = paste0("FeaturePlot_", cluster, ".png"))
  
  # Save the plot
  ggsave(filename, plot = gg, width = 6, height = 5, dpi = 500)
  print(gg)
  
  rm(gg)
  gc()
}

```


### Save RDS
```{r}
myfile <- file.path(dir_obj, paste0("06_", TISSUE, "_IntestinalSeurat_TissueSubsetReintegrated_ModuleScores.rds"))
myfile

saveRDS(obj, file = myfile)
```



***

# Clustering: SI-IEL-specific clustering and figure preparation

The following analyses reproduce the SI-IEL-specific clustering and annotations used for manuscript Figure 3.

Seurat cluster is overridden with manual cluster order.

```{r}
stopifnot(TISSUE == "SI-IEL")
```

## Load RDS
```{r}
# Frozen checkpoint for SI-IEL figure preparation
if (use_frozen_object) {
  frozen_obj_file <- file.path(dir_frozen, "06_SI-IEL_IntestinalSeurat_TissueSubsetReintegrated_ModuleScores_FROZEN.rds")
  
  obj <- LoadSeuratRds(frozen_obj_file)
  
  message("Using frozen SI-IEL checkpoint for exact Figure reproduction.")
}
```

```{r}
obj
# An object of class Seurat 
# 39647 features across 6071 samples within 5 assays 
# Active assay: SCT (15709 features, 0 variable features)
#  3 layers present: counts, data, scale.data
#  4 other assays present: RNA, HTO, CITE, integrated
#  2 dimensional reductions calculated: pca, umap

# colnames(obj@meta.data)
```


```{r}
##### 1) Clustering on integrated assay
# set integrated assay
DefaultAssay(obj) <- "integrated"

# Run PCA and clustering
obj <- RunPCA(
  obj,
  assay   = "integrated",
  features = VariableFeatures(obj, assay = "integrated")
)

obj <- FindNeighbors(obj, dims = 1:numberDimsPCA_for_UMAP)
obj <- FindClusters(obj, resolution = res_for_clustering) # parameter section, was set to 0.5

# keep genotype ordering
# Ensure that the order is WT on the left and the renamed KO on the right
obj$genotype <- factor(obj$genotype, levels = c("WT", "KO"))

###
# Create a new vector for desired cluster order (swapping 7 and 8)
new_cluster_ids <- c(0, 1, 2, 3, 4, 5, 6, 8, 7, 9, 10, 11, 12)

# # Assign new identities (overwrite seurat_clusters)
# # and reorder clusters to 0:12
obj$seurat_clusters <- plyr::mapvalues(
  x    = obj$seurat_clusters,
  from = levels(obj$seurat_clusters), #original seurat clusters
  to   = new_cluster_ids
)

# Update seurat_clusters levels to be continuous 0 to 12
obj$seurat_clusters <- factor(
  obj$seurat_clusters,
  levels = 0:12
)

# Visualize the UMAP plot with the new continuous cluster labels
p1 <- DimPlot(obj,
        reduction = "umap",
        split.by = "genotype",
        group.by = "seurat_clusters",
        label = TRUE) +
  theme(strip.text = element_text(face = "bold"))

print(p1)

# Save the first UMAP plot
p1file <- make_path(dir = dir_plots, filename = "UMAP_seurat_clusters.svg")
ggsave(filename = p1file, plot = p1, width = 10, height = 6, device = "svg")


######
# 4) Annotate clusters manually
# Create a new metadata column called 'manual_annotation'
# same as fig
cluster_annotations <- c(
  "0 Trm CD4-CD8αα+",            # 0
  "1 Trm CD4-CD8α+βlow",        # 1
  "2 Trm CD4-CD8αα+",            # 2
  "3 Trm CD4-CD8α+βlow",        # 3
  "4 Tnaive CD4+ & CD8αβ+",      # 4
  "5 Tfh (CD4+CD8-)",            # 5
  "6 rmCTL CD4-CD8α+βlow",      # 6
  "7 rmCTL CD4-CD8αα+",          # 7
  "8 Tnaive CD4+ & CD8αβ+",      # 8
  "9 Treg (CD4+CD8-)",           # 9
  "10 Tprol CD8αβ+",              # 10
  "11 Not assigned",              # 11
  "12 Not assigned"               # 12
)

# Map annotations to clusters (adjust for 13 clusters)
obj$manual_annotation <- factor(
  obj$seurat_clusters,
  levels = 0:12,
  labels = cluster_annotations
)

######
# Visualize UMAP with manual annotations
p2 <- DimPlot(obj, reduction = "umap", group.by = "manual_annotation", label = TRUE)

print(p2)

# Save the second UMAP plot
p2file <- make_path(dir = dir_plots, filename = "UMAP_manual_annotations.svg")
ggsave(p2file, plot = p2, width = 10, height = 6, device = "svg")

```



## Count cells (Alluvial plot) Fig3F

Count cells in each cluster for WT and KO and do the alluvial plot

```{r}
# Count cells in each cluster for each genotype (WT and KO)
cell_counts <- obj@meta.data %>%
  group_by(genotype, seurat_clusters) %>%
  summarise(count = n()) %>%
  as.data.frame()

# Calculate the percentage of each cluster within each genotype
cell_counts <- cell_counts %>%
  group_by(genotype) %>%
  mutate(percentage = (count / sum(count)) * 100,
         Percentage_Label = paste0(count, " (", round(percentage, 1), "%)")) %>%
  ungroup()

# Rename columns to match ggalluvial structure
# Var1 = genotype, Var2 = cluster
plt1 <- cell_counts %>%
  rename(Var1 = genotype,
         Var2 = seurat_clusters,
         Percent = percentage)

# Create alluvial plot
alluvial_plot <- ggplot(plt1,
                        aes(x = Var1, #genotype
                            stratum = Var2, #cluster
                            alluvium = Var2, #same cluster across genotypes
                            y = Percent,
                            fill = factor(Var2))) +
  geom_flow(stat = "alluvium", lode.guidance = "forward", 
            alpha = 0.7, size = 0.1, colour = NA) +
  geom_stratum(alpha = 0.9, width = 0.6, 
               size = 0.1, colour = "black") +
  geom_text(aes(label = Percentage_Label),
            stat = "stratum", size = 4, colour = "black") +
  xlab("") +
  ylab("Percentage of cells") +
  labs(title = "Cluster Distribution Changes from WT to KO",
       fill = "Cluster") +
  theme_classic(base_size = 25) +
  theme(legend.title = element_text(size = 14),
        legend.text = element_text(size = 12),
        plot.title = element_text(size = 15, face = "bold"))

# Print plot
print(alluvial_plot)

# Save plot
filename <- make_path(dir = dir_plots, filename = "genotype_alluvialplot.svg")
ggsave(filename = filename,
       plot = alluvial_plot, width = 12, height = 10, device = "svg")


```


## ViolinPlots

Make vln plot combine CITE-seq and markers (look saved file _ Figure 3E)

```{r}
writeLines("Dimension reduction colored by CITE-seq protein abundance:")

# --- Set up assays and marker lists ---
DefaultAssay(obj) <- "CITE"

# List of specific CITE-seq markers to visualize
cite_markers <- c("CD4", "CD8a", "CD8b")

# List of additional genes to visualize
rna_markers <- c("Itgae", "Runx3",  "Cd44", "Sell", "Cxcr5", "Gzmb", "Foxp3", "Tbx21", "Mki67")

# Combine the CITE-seq markers and RNA markers
all_markers <- unique(c(cite_markers, rna_markers))

# Create a list to store the violin plots
vln_plots <- list()

# --- Generate violin plots for each marker ---
for (i in all_markers) {
  
  # Switch assay automatically
  if (i %in% cite_markers) {
    DefaultAssay(obj) <- "CITE"
  } else {
    DefaultAssay(obj) <- "RNA"
  }
  
  # Check if marker exists in data
  if (i %in% rownames(GetAssayData(obj, slot = "data"))) {
    
    # Create the violin plot
    vln_plot <- VlnPlot(
      obj, 
      features = i, 
      group.by = "seurat_clusters",
      pt.size = 0,
      slot = "data"
    ) +
       scale_y_continuous(limits = c(0, 4), breaks = c(0, 2, 4)) +  # <-- fix y-axis range
      labs(x = "", y = "") +
      theme(
        plot.title = element_blank(),
        axis.text.x = element_text(size = 25, angle = 0, hjust = 0.5),
        axis.text.y = element_text(size = 15),
        plot.margin = unit(c(1, 2, 1, 2), "cm")  # extra space on left
      ) +
      coord_cartesian(clip = "off")  # allow text outside plot area
    
    # Add title OUTSIDE plot area, to the left of y-axis
    vln_plot <- vln_plot +
      annotation_custom(
        grob = textGrob(
          label = i,
          gp = gpar(fontsize = 25),
          x = unit(-0.06, "npc"),   # position outside plot (-0.1 = 10% left)
          y = unit(0.5, "npc"),    # vertically centered
          just = "right"
        )
      )
    
    # Store plot without legend
    vln_plots[[i]] <- vln_plot + NoLegend()
    
  } else {
    print(paste0(i, " not found in data"))
  }
}

# --- Combine all violin plots into one vertical panel ---

combined_vln_plot <- wrap_plots(vln_plots, ncol = 1) +
  plot_layout(heights = rep(1, length(vln_plots))) &
  theme(plot.margin = margin(0, 0, 0, 2, "cm"))

# Display combined plot
print(combined_vln_plot)

# --- Save to PDF ---
filename.svg <- make_path(dir = dir_plots, filename = "combined_violin_plots_CITE_RNA.svg")
filename.png <- make_path(dir = dir_plots, filename = "combined_violin_plots_CITE_RNA.png")


ggsave(
  filename = filename.svg,
  plot = combined_vln_plot,
  width = 11.5, height = 13,
  limitsize = FALSE, device = "svg"
)

ggsave(
  filename = filename.png,
  plot = combined_vln_plot,
  width = 11.5, height = 13,
  limitsize = FALSE, device = "png"
)


```


## Runx3 ViolinPlot 

```{r}
DefaultAssay(obj) <- "RNA"

target_gene <- "Runx3"

# check
if (!target_gene %in% rownames(GetAssayData(obj, slot = "data"))) {
  stop("Runx3 not found in RNA assay")
}

p_runx3 <- VlnPlot(
  obj,
  features = target_gene,
  group.by = "seurat_clusters",
  pt.size = 0.5,
  slot = "data"
) +
  labs(x = "", y = "") +
  ggtitle("Runx3 log normalized gene expression") +
  theme(
    plot.title = element_text(size = 30, hjust = 0.5),
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 15)
  ) +
  NoLegend()

# save 
save_plot(
  plot = p_runx3,
  dir  = dir_plots,
  name = "Runx3_violin_plot",
  w = 8,
  h = 6
)

```



## Marker Genes for Clusters 

Using Seurat’s FindMarkers function, we can test for gene-wise differential expression between a specific cell cluster and its respective complement (i.e. all cells from all other clusters). The DE-testing results are filtered for positive fold changes (i.e. gene is up in the respective cluster) and the expression pattern of the top 30 genes per test are illustrated by heatmap. The full table results for each cluster, containing all significant genes with positive fold changes per cluster, are stored in the "Results" directory.

```{r}
## Set identification and extract loop variables (the clusters)
Idents(obj) <- "seurat_clusters"
loop_vars <- Idents(obj) %>% levels() 

# list to save markers per cluster
top_markers_list <- list()
complete_markers_list <- list()

#### 
## FindMarkers
DefaultAssay(obj) <- "SCT"
# find markers that define tissue types using FindMarkers function on the SCT assay
obj <- PrepSCTFindMarkers(obj)

## Perform loop (differential analysis per cluster vs rest)
for (i in loop_vars){
  
  # print
  message(paste("FindMarkers of Cluster",i))
  
  # get Markers from this cluster only 
  resultFindMarkers <- FindMarkers(obj, assay = "SCT", ident.1 = i)
  
  # subselect for positive Fold change
  # if there are any
  if (nrow(resultFindMarkers) > 0) {
    resultFindMarkers <- resultFindMarkers %>% filter(avg_log2FC > 0 &  p_val_adj < 0.05)
  }
  
  # gene names as new column
  GeneNames <- resultFindMarkers %>% rownames()
  resultFindMarkers <- cbind(GeneNames, resultFindMarkers)

  # get manual annotation label for this cluster i
  # will repeat for every row
  this_annot <- obj@meta.data %>%
    as.data.frame() %>%
    dplyr::filter(seurat_clusters == i) %>%
    dplyr::pull(manual_annotation) %>%
    unique()
  
  # add to table
  resultFindMarkers$manual_annotation <- this_annot
  resultFindMarkers <- resultFindMarkers %>%
  dplyr::rename(Cluster = manual_annotation) #new name = old name
  
  # select top results (for subsequent heatmap plotting)
  # if there are any significant
  if (nrow(resultFindMarkers) > 0) {
    resultFindMarkers_top30 <- resultFindMarkers %>%
        slice_head(n = 30) #get top 30 even when n < 30 and n = 0.
  
    # save markers and top30 to list
    complete_markers_list[[i]] <- resultFindMarkers
    
    top_markers_list[[i]] <- resultFindMarkers_top30
    
  }
  
# save result as table
  filename_table <- make_path(dir = dir_tables, filename = paste0("MarkerGenes_padj005_Cluster_", i,".txt"))

  write.table(x = resultFindMarkers, 
              file = filename_table,
              col.names =TRUE,
              row.names = FALSE, 
              quote=FALSE, 
              sep="\t")
  
}



```


## Heatmap for showing all genes unique for each cluster (figure 3B)
```{r}
# Set identity and assay
Idents(obj) <- "seurat_clusters"
DefaultAssay(obj) <- "SCT"

# Find all markers
hall_markers <- FindAllMarkers(
  obj,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

# Filter markers for significant padj005
hall_markers_filtered <- hall_markers %>%
  filter(avg_log2FC > 0, p_val_adj < 0.05)

# Select top markers per cluster
htop_markers <- hall_markers_filtered %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 20) %>% ##### top n
  ungroup()

# Get list of unique top genes
heatmap_features <- unique(htop_markers$gene)


# Compute average expression per cluster
avg_expr <- AverageExpression(
  obj,
  features = heatmap_features,
  group.by = "seurat_clusters",
  assays = "SCT"
)$SCT



# set as matrix
avg_expr <- as.matrix(avg_expr)

# scale across genes (rows)
scaled_expr <- t(scale(t(avg_expr)))  # Z-score across clusters

# Convert to long format for ggplot
heatmap_df <- as.data.frame(scaled_expr) %>%
  mutate(gene = rownames(.)) %>%
  pivot_longer(-gene, names_to = "cluster", values_to = "expression")

# Fix cluster order from 0 to 12
cluster_levels <- as.character(sort(as.numeric(unique(heatmap_df$cluster))))
# apply it 
heatmap_df$cluster <- factor(heatmap_df$cluster, 
                             levels = cluster_levels)
# gene as factor 
heatmap_df$gene <- factor(heatmap_df$gene,
                          levels = rev(unique(heatmap_df$gene)))
# Keep gene order top-down

# Plot with ggplot2 (mimicking DoHeatmap style)
heatmap_plot <- ggplot(heatmap_df,
                       aes(x = cluster,
                           y = gene, 
                           fill = expression)) +
  geom_tile(color = NA) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Expression") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 10),
    axis.title = element_blank(),
    legend.position = "left",
    panel.grid = element_blank()
  ) +
  labs(title = "Average Scaled Expression of Top Markers per Cluster")

# Save plot
filename <- make_path(dir = dir_plots, filename = "Fig3_FindMarkers_padj005_top20_Heatmap.pdf")
filename

ggsave(filename = filename,
       plot = heatmap_plot,
       width = 12, height = 20)

# Print to screen
print(heatmap_plot)

```


## Heatmap of the top 30

```{r}

# Select top markers per cluster
htop_markers <- hall_markers_filtered %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 30) %>% ##### top n
  ungroup()

# Get list of unique top genes
heatmap_features <- unique(htop_markers$gene)


# Compute average expression per cluster
avg_expr <- AverageExpression(
  obj,
  features = heatmap_features,
  group.by = "seurat_clusters",
  assays = "SCT"
)$SCT



# set as matrix
avg_expr <- as.matrix(avg_expr)

# scale across genes (rows)
scaled_expr <- t(scale(t(avg_expr)))  # Z-score across clusters

# Convert to long format for ggplot
heatmap_df <- as.data.frame(scaled_expr) %>%
  mutate(gene = rownames(.)) %>%
  pivot_longer(-gene, names_to = "cluster", values_to = "expression")

# Fix cluster order from 0 to 12
cluster_levels <- as.character(sort(as.numeric(unique(heatmap_df$cluster))))
# apply it 
heatmap_df$cluster <- factor(heatmap_df$cluster, 
                             levels = cluster_levels)
# gene as factor 
heatmap_df$gene <- factor(heatmap_df$gene,
                          levels = rev(unique(heatmap_df$gene)))
# Keep gene order top-down

# Plot with ggplot2 (mimicking DoHeatmap style)
heatmap_plot <- ggplot(heatmap_df,
                       aes(x = cluster,
                           y = gene, 
                           fill = expression)) +
  geom_tile(color = NA) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Expression") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 10),
    axis.title = element_blank(),
    legend.position = "left",
    panel.grid = element_blank()
  ) +
  labs(title = "Average Scaled Expression of Top Markers per Cluster")

# Save plot
filename <- make_path(dir = dir_plots, filename = "Fig3_FindMarkers_padj005_top30_Heatmap.pdf")
filename

ggsave(filename = filename,
       plot = heatmap_plot,
       width = 12, height = 20)

# Print to screen
print(heatmap_plot)


```


## Heatmap of all Markers

```{r}

# Select top markers per cluster
htop_markers <- hall_markers_filtered %>%
  group_by(cluster) %>%
  arrange(desc(avg_log2FC), .by_group = TRUE) %>% #arrange but no slice
  ungroup()

# Get list of unique top genes
heatmap_features <- unique(htop_markers$gene)


# Compute average expression per cluster
avg_expr <- AverageExpression(
  obj,
  features = heatmap_features,
  group.by = "seurat_clusters",
  assays = "SCT"
)$SCT



# set as matrix
avg_expr <- as.matrix(avg_expr)

# scale across genes (rows)
scaled_expr <- t(scale(t(avg_expr)))  # Z-score across clusters

# Convert to long format for ggplot
heatmap_df <- as.data.frame(scaled_expr) %>%
  mutate(gene = rownames(.)) %>%
  pivot_longer(-gene, names_to = "cluster", values_to = "expression")

# Fix cluster order from 0 to 12
cluster_levels <- as.character(sort(as.numeric(unique(heatmap_df$cluster))))
# apply it 
heatmap_df$cluster <- factor(heatmap_df$cluster, 
                             levels = cluster_levels)
# gene as factor 
heatmap_df$gene <- factor(heatmap_df$gene,
                          levels = rev(unique(heatmap_df$gene)))
# Keep gene order top-down

# Plot with ggplot2 (mimicking DoHeatmap style)
heatmap_plot <- ggplot(heatmap_df,
                       aes(x = cluster,
                           y = gene, 
                           fill = expression)) +
  geom_tile(color = NA) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "Expression") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12, angle = 0, hjust = 1),
    axis.text.y = element_text(size = 10),
    axis.title = element_blank(),
    legend.position = "left",
    panel.grid = element_blank()
  ) +
  labs(title = "Average Scaled Expression of all Markers per Cluster")

# Save plot
filename <- make_path(dir = dir_plots, filename = "Fig3_FindMarkers_padj005_allmarkers_Heatmap.pdf")
filename

ggsave(filename = filename,
plot = heatmap_plot,
width = 12, height = 20)

# Print to screen
print(heatmap_plot)
```


# Save RDS

```{r}

filename <- file.path(dir_obj, paste0("06_", TISSUE, "_IntestinalSeurat_integrated_ModuleScoresManualAnnot.rds"))

saveRDS(obj, filename)

```

