---
title: "HDAC1 and HDAC2 Intestinal Immunity - Transcription Factor Activity Inference"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
---

# Transcription Factor Activity Inference

Here we do:

- TF activity inference with DecoupleR

The intestinal tissue subsets can be three different types:

1. Small intestinal intraepithelial **(SI-IEL)**

2. Small intestinal lamina propria **(SI-LPL)**

3. Colon

The previous script `06_Intestinal_scRNAseq_ModuleScores_MarkerGenes.Rmd` produces the RDS object that is necessary for this downstream analysis.

Here it is possible to select one out of the three tissues and run all downstream analysis. See the *parameters* section.

# Setup 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(42)
```

## Libraries
```{r, message=FALSE, warnings=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(RColorBrewer)
library(knitr)
library(tibble)

library(Seurat)
library(SeuratData)

library(pheatmap)
library(decoupleR)
library(OmnipathR)
```

## TISSUE
```{r}
TISSUE <- "SI-IEL"

stopifnot(TISSUE %in% c("SI-IEL", "SI-LPL", "colon"))
```

## Object source

By default, manuscript figures are reproduced using the frozen Seurat object used for the final analysis. Set `use_frozen_object = FALSE` to use the object generated by the preceding pipeline scripts.

*Minor differences in stochastic dimensionality-reduction results may occur when the full workflow is rerun from raw data.*

```{r}
use_frozen_object <- TRUE
```


## Directories
```{r}
dir_input <- here("Data", "Processed", "06_ModuleScores", TISSUE)
dir_frozen <- here("Data", "Processed", "00_Frozen")

dir_res <- here("Results", "07_TFactivity", TISSUE) #parent dir
dir_obj <- here("Data", "Processed", "07_TFactivity", TISSUE) # objects
dir_plots <- file.path(dir_res, "Plots")
dir_tables <- file.path(dir_res, "Tables")

dirs <- c(dir_obj, dir_plots, dir_tables)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```


## Genotype mapping
```{r}
tissue_dict <- list(
  "SI-IEL" = c("WT-SI-IEL", "KO-SI-IEL"),
  "SI-LPL" = c("WT-SI-LPL", "KO-SI-LPL"),
  "colon"  = c("WT-colon",  "KO-colon")
)


subselection_HTOclasses <- tissue_dict[[TISSUE]]

ko <- subselection_HTOclasses[grepl("^KO", subselection_HTOclasses)]
wt <- subselection_HTOclasses[grepl("^WT", subselection_HTOclasses)]

DEgroups_HTOclasses <- list(list(ko, wt))

message("Selected tissue: ", TISSUE)
message("WT/KO groups: "); print(subselection_HTOclasses)
message("Pairwise DE comparison: "); print(DEgroups_HTOclasses)
```

## Helper functions

```{r}
# make path
make_path <- function(dir, filename) {
  file.path(dir, paste0(TISSUE, "_", filename))
}

# save all extensions for the plots
save_plot <- function(plot, dir, name, w = 10, h = 6) {
  base <- file.path(dir, paste0(TISSUE, "_", name))
  ggsave(paste0(base, ".pdf"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".png"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".svg"), plot = plot, width = w, height = h)
}


```

# Load data
```{r}
frozen_obj_file <- file.path(
  dir_frozen,
  "06_SI-IEL_IntestinalSeurat_integrated_ModuleScoresManualAnnot_FROZEN.rds"
)

obj_file <- file.path(
  dir_input,
  paste0("06_", TISSUE, "_IntestinalSeurat_integrated_ModuleScoresManualAnnot.rds")
)

if (use_frozen_object) {
  stopifnot(TISSUE == "SI-IEL")

  obj <- readRDS(frozen_obj_file)
  message("Using frozen Seurat object for exact manuscript figure reproduction.")
  
} else {
  obj <- readRDS(obj_file)
  message("Using Seurat object generated by the reproducible analysis pipeline.")

  }
```


## Check
```{r}
DefaultAssay(obj) <- "integrated"

# check clusters and annotations exist
table(obj$seurat_clusters)
table(obj$manual_annotation)

# check UMAP exists
if (!"umap" %in% names(obj@reductions)) {
  stop("UMAP is missing — you must run UMAP in integration script, not here.")
}

p1 <- DimPlot(obj,
        reduction = "umap",
        split.by = "genotype",
        group.by = "seurat_clusters",
        label = TRUE) +
  theme(strip.text = element_text(face = "bold"))

p1

```

## Manual annotation
```{r}
cluster_annotations <- c(
  "0 Trm CD4-CD8αα+",            # 0
  "1 Trm CD4-CD8α+βlow",        # 1
  "2 Trm CD4-CD8αα+",            # 2
  "3 Trm CD4-CD8α+βlow",        # 3
  "4 Tnaive CD4+ & CD8αβ+",      # 4
  "5 Tfh (CD4+CD8-)",            # 5
  "6 rmCTL CD4-CD8α+βlow",      # 6
  "7 rmCTL CD4-CD8αα+",          # 7
  "8 Tnaive CD4+ & CD8αβ+",      # 8
  "9 Treg (CD4+CD8-)",           # 9
  "10 Tprol CD8αβ+",              # 10
  "11 Not assigned",              # 11
  "12 Not assigned"               # 12
)

# Map annotations to clusters
obj$manual_annotation <- factor(
  obj$seurat_clusters,
  levels = 0:12,
  labels = cluster_annotations
)

######
# Visualize UMAP with manual annotations
p2 <- DimPlot(obj, reduction = "umap", group.by = "manual_annotation", label = TRUE)

print(p2)
```


***

# CollecTri 

Here we will do a Transcription Factor Activity Inference Analysis using decoupleR's ULm algortihm with the Collectri TF database.

ULM means **Univariate Linear Model**. *ULM fits a linear model for each sample and regulator, where the observed molecular readouts in mat are the response variable and the regulator weights in net are the explanatory one. Target features with no associated weight are set to zero. The obtained t-value from the fitted model is the activity ulm of a given regulator.* [https://saezlab.github.io/decoupleR/reference/run_ulm.html]

For each TF, ULM looks at its target genes, uses mode of regulation (mor, +1/−1), and estimates a TF activity score per cell. So the TF activity is the weighted behavior of its targets

* Source (The TF)
* Target (TF's target gene)
* MOR: Mode of Regulation (+1 activates the target, -1 represses the target)

Thus, if TF is active:  
activated targets (+1) go up  
repressed targets (-1) go down  


```{r}
# get normalized log-transformed counts
mat <- GetAssayData(object = obj, layer = "data", assay = "RNA")

if (use_frozen_object) {
  stopifnot(TISSUE == "SI-IEL")
  
  frozen_acts_file <- file.path(dir_frozen,"07_SI-IEL_IntestinalSeurat_Collectri_Mouse_ULM_activities_FROZEN.rds")
  acts <- readRDS(frozen_acts_file)
  message("Using frozen ULM activities for exact manuscript figure reproduction.")
  
  
} else { # get and run Collectri now
  
  net <- get_collectri(organism='mouse', split_complexes=FALSE)
  # [2025-12-15 08:11:58] [SUCCESS] [OmnipathR] Downloaded 40291 interactions.
  net
  # # A tibble: 39,961 × 3
  #    source target   mor
  #    <chr>  <chr>  <dbl>
  #  1 Myc    Tert       1
  #  2 Spi1   Bglap2     1
  #  3 Spi1   Bglap      1
  #  4 Spi1   Bglap3     1
  #  5 Smad3  Jun        1
  #  6 Smad4  Jun        1
  #  7 Stat5a Il2        1
  #  8 Stat5b Il2        1
  #  9 Rela   Fas        1
  # 10 F6Y2T6 Nr0b1      1
  # # ℹ 39,951 more rows
  
  
  # check compatibility between RNA features and CollecTRI targets
  length(intersect(rownames(mat), net$target))
  length(rownames(mat))
  
  # run ULM
  acts <- decoupleR::run_ulm(
    mat = mat,
    net = net,
    .source = "source",
    .target = "target",
    .mor = "mor",
    minsize = 5
  ) 
  
  acts_file <- file.path(dir_obj, paste0("07_", TISSUE, "_IntestinalSeurat_Collectri_Mouse_ULM_activities.rds"))
  
  saveRDS(acts, acts_file)
  message("ULM activities generated by the reproducible analysis pipeline.")
}
```


## Check
```{r}
acts
# # A tibble: 3,788,304 × 5
# # ℹ 3,788,294 more rows

```


## Store in obj (tfsulm assay)
```{r}
# Extract ulm and store it in tfsulm in obj
obj[['tfsulm']] <- acts %>%
  pivot_wider(id_cols = 'source' ,
              names_from = 'condition',
              values_from = 'score') %>%
  column_to_rownames('source') %>%
  Seurat::CreateAssayObject(.)
```

```{r}
# change assay
DefaultAssay(obj) <- "tfsulm"

# scale data
# store it in specific assay, not overwritting scale data
obj <- ScaleData(obj, assay = "tfsulm")

obj
# An object of class Seurat 
# 40271 features across 6071 samples within 6 assays 
# Active assay: tfsulm (624 features, 0 variable features)
#  3 layers present: counts, data, scale.data
#  5 other assays present: RNA, HTO, CITE, SCT, integrated
#  2 dimensional reductions calculated: pca, umap

# copy scaled ULM activities to the data slot because downstream plotting and differential-activity code uses this slot
obj[["tfsulm"]]@data <- obj[["tfsulm"]]@scale.data
```


```{r warning=FALSE}
all.equal(
  GetAssayData(obj, assay = "tfsulm", slot = "data"),
  GetAssayData(obj, assay = "tfsulm", slot = "scale.data")
)
# should be true
```


# Visualize 

## Gene of Interest

```{r}
gene_of_interest <- c("Zbtb7b")
```

## Setup
```{r}
DefaultAssay(obj) <- "tfsulm"
DefaultAssay(obj) # needs to be tfsulm

# check whether the gene of interest is actually present
gene_exists <- function(gene, object, assay) {
  gene %in% rownames(object[[assay]])
}

```

## TF Activity and Expression (Fig 4C)

Where in the UMAP do cells show high inferred activity of this TF gene of interest?

Red > 0 TF is more active than average in these cells  
white = 0 TF activity near global average  
Blue < 0 TF is less active than average  
  
```{r warning=FALSE}
for (gene in gene_of_interest) {
  
  if (!gene_exists(gene, obj, "tfsulm")) {
    message("Skipping ", gene, " – not found in tfsulm assay")
    next
  }
  
  p_tf <- FeaturePlot(
    obj,
    features = gene,
    reduction = "umap",
    slot = "scale.data",
  ) +
    scale_colour_gradient2(
      low = "darkblue",
      mid = "white",
      high = "red"
    ) +
    ggtitle(paste(gene, "TF Activity (ULM, scaled)"))
  
  save_plot(plot = p_tf, # tf activity
            dir = dir_plots,
            name = paste0("TF_activity_", gene),
            w = 8, h = 6
            )
  
  ## with labels
  Idents(obj) <- obj@meta.data$manual_annotation
  
  p_tf2 <- FeaturePlot(
    obj,
    features = gene,
    reduction = "umap",
    slot = "scale.data",
    label = TRUE
  ) +
    scale_colour_gradient2(
      low = "darkblue",
      mid = "white",
      high = "red"
    ) +
    ggtitle(paste(gene, "TF Activity (ULM, scaled)"))
  
  save_plot(plot = p_tf2, # tf activity
            dir = dir_plots,
            name = paste0("TF_activity_manual_annotation_", gene),
            w = 8, h = 6
            )
  
  ## with clusters
  Idents(obj) <- obj@meta.data$seurat_clusters
  
  p_tf3 <- FeaturePlot(
    obj,
    features = gene,
    reduction = "umap",
    slot = "scale.data",
    label = TRUE
  ) +
    scale_colour_gradient2(
      low = "darkblue",
      mid = "white",
      high = "red"
    ) +
    ggtitle(paste(gene, "TF Activity (ULM, scaled)"))
  
  save_plot(plot = p_tf3, # tf activity
            dir = dir_plots,
            name = paste0("TF_activity_seurat_clusters_", gene),
            w = 8, h = 6
            )
  
  #####
  ### TF expression
  DefaultAssay(object = obj) <- "RNA"

  p_exp <- FeaturePlot(obj,
                       features = gene,
                       reduction = "umap",
                       slot = "data") + ggtitle(paste(gene, "Expression (RNA, log-normalized counts)"))
  
  save_plot(plot = p_exp, # tf expression
            dir = dir_plots,
            name = paste0("TF_expression_", gene),
            w = 8, h = 6
            )
  
  # set it back to tfsulm
  DefaultAssay(object = obj) <- "tfsulm"
  
  ## print plots
p_tf | p_exp
p_tf2 | p_tf3
}


```



*** 

# TF activity difference between KO vs WT 

Cell-level comparison of inferred TF activity between KO and WT, ignoring clusters
Here we will answer **which TFs change with genotype (HDAC KO)**

Test:
- t-test

It is a *z-scored TF activities*

delta_difference > 0 --> TF activity higher in KO
delta_difference < 0--> TF activity higher in WT

## Setup
```{r}
DefaultAssay(obj) <- "tfsulm"

# get the tfs
mat <- GetAssayData(obj, assay = "tfsulm", slot = "data")
tfs <- rownames(mat)

# get groups
group <- obj$genotype
# factor relevel 
group <- factor(group, levels = c("KO", "WT"))
levels(group)

```

## Per-TF differential test
```{r warning=FALSE}

res_list <- lapply(tfs, function(tf) {
  # get matrix values - the TF activity scores
  vals <- as.numeric(mat[tf, ]) 
  # groups 
  ko_vals <- vals[group == "KO"]
  wt_vals <- vals[group == "WT"]
  
  # perform t-test
  test <- t.test(ko_vals, wt_vals) # default is Welch
  
  # mean and delta 
  mean_KO <- mean(ko_vals, na.rm = TRUE)
  mean_WT <- mean(wt_vals, na.rm = TRUE)
  
  # save res
  data.frame(
    TF = tf,
    mean_KO = mean_KO, 
    mean_WT = mean_WT,
    delta_activity = mean_KO - mean_WT,
    p_val = test$p.value,
    stringsAsFactors = FALSE
  )
})


## bind results of test
tf_diff <- do.call(rbind, res_list)
# correct pvals
tf_diff$p_val_adj <- p.adjust(tf_diff$p_val, method = "BH")

tf_diff <- tf_diff %>%
  arrange(p_val_adj)


# filter significant tfs
tf_diff_padj005 <- tf_diff %>% dplyr::filter(p_val_adj < 0.05)
tf_diff_padj005 <- tf_diff_padj005 %>%
  arrange(p_val_adj)

# save res
## all
filename.all <- make_path(dir = dir_tables, filename = "TF_activity_KOvsWT_all.csv")
write.csv(tf_diff, file = filename.all, row.names = FALSE)

## padj 005
filename005 <- make_path(dir = dir_tables, filename = "TF_activity_KOvsWT_padj005.csv")
write.csv(tf_diff_padj005, file = filename005, row.names = FALSE)

```

## Summary

```{r}
# total TFs tested
n_total <- nrow(tf_diff)

# p-value related issues
n_na_pval <- sum(is.na(tf_diff$p_val))
n_na_padj <- sum(is.na(tf_diff$p_val_adj))

# significant TFs
n_sig_padj005 <- sum(tf_diff$p_val_adj < 0.05, na.rm = TRUE)


# print
cat(
  "TF differential activity QC summary\n",
  "----------------------------------\n",
  "Total TFs tested:              ", n_total, "\n",
  "NA p-values:                   ", n_na_pval, "\n",
  "NA adjusted p-values:          ", n_na_padj, "\n",
  "Significant TFs (padj < 0.05): ", n_sig_padj005, "\n",
  sep = ""
)

# TF differential activity QC summary
# ----------------------------------
# Total TFs tested:              624
# NA p-values:                   0
# NA adjusted p-values:          0
# Significant TFs (padj < 0.05): 487

```

***

# Top 100 TFs heatmap

We will answer **which TFs best distinguish CD4 T cell clusters** here:

Here we do:  
1. get the TF activity per cell  
2. gets the mean TF activity per cluster  
3. select the top 100 variable TFs across clusters
4. Plot a heatmap of TF per cluster

## Setup
```{r}
# idents are seurat clusters
Idents(obj) <- "seurat_clusters"
table(Idents(obj))

# how many top TFs to get
n_tfs <- 100

# get ULM assay
DefaultAssay(obj) <- "tfsulm"

# get TF activity matrx
# Note: tfsulm@data contains scaled ULM activities
# (scale.data was intentionally copied into data earlier)
mat <- GetAssayData(obj, assay = "tfsulm", slot = "data")

```


## Heatmap

Legend is Zscored TF activity computed with ULM:  

* Blue: low relative TF activity  

* White: average activity

* Red: high relative TF activity


```{r}
## df pivot
# long: cell -> cluster -> TF
df_tf <- t(as.matrix(mat)) %>% #transpose so cells are rows 
  as.data.frame() %>%
  dplyr::mutate(cluster = Idents(obj)) %>%
  tidyr::pivot_longer( #long df
    cols = -cluster,
    names_to = "TF",
    values_to = "activity"
  ) %>%
  dplyr::group_by(cluster, TF) %>%
  dplyr::summarise(mean_activity = mean(activity), .groups = "drop")

# get top TFs by variability across clusters
top_tfs <- df_tf %>%
  dplyr::group_by(TF) %>%
  dplyr::summarise(sd_across_clusters = sd(mean_activity), .groups = "drop") %>%
  dplyr::arrange(desc(abs(sd_across_clusters))) %>%
  dplyr::slice_head(n = n_tfs) %>%
  dplyr::pull(TF) #only th TFs

# build matrix: TF vs cluster
top_acts_mat <- df_tf %>%
  dplyr::filter(TF %in% top_tfs) %>% #filter for only the top tfs selected before
  tidyr::pivot_wider(
    id_cols = cluster,
    names_from = TF,
    values_from = mean_activity
  ) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix() %>%
  t() #tanspose

# check range
range(top_acts_mat, na.rm = TRUE)
#[1] -3.003488  3.406092

# color scale, dynamic, based on data
palette_length <- 100
my_color <- colorRampPalette(c("blue", "white", "red"))(palette_length)

# heatmap breaks set
my_breaks <- c(
  seq(-3, 0, length.out = ceiling(palette_length / 2) + 1),
  seq(0.05, 3, length.out = floor(palette_length / 2))
)

####
###
# Heatmap

# file
heatmap_base <- make_path(dir = dir_plots, filename = paste0("TF_activity_heatmap_SeuratClusters_top", n_tfs))

## PNG
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = TRUE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 0, # dont rotate cluster labels here
  filename = paste0(heatmap_base, ".png"),
  width = 6,
  height = 0.15 * n_tfs
)

## pdf
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = TRUE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 0, # dont rotate cluster labels here
  filename = paste0(heatmap_base, ".pdf"),
  width = 6,
  height = 0.15 * n_tfs
)

```


## Manual annotation 
```{r}
table(obj$manual_annotation)
# idents manual annot
Idents(obj) <- "manual_annotation"
table(Idents(obj))


```

```{r}
## df pivot
# long: cell -> cluster -> TF
df_tf <- t(as.matrix(mat)) %>% #transpose so cells are rows 
  as.data.frame() %>%
  dplyr::mutate(cluster = Idents(obj)) %>%
  tidyr::pivot_longer( #long df
    cols = -cluster,
    names_to = "TF",
    values_to = "activity"
  ) %>%
  dplyr::group_by(cluster, TF) %>%
  dplyr::summarise(mean_activity = mean(activity), .groups = "drop")

# get top TFs by variability across clusters
top_tfs <- df_tf %>%
  dplyr::group_by(TF) %>%
  dplyr::summarise(sd_across_clusters = sd(mean_activity), .groups = "drop") %>%
  dplyr::arrange(desc(abs(sd_across_clusters))) %>%
  dplyr::slice_head(n = n_tfs) %>%
  dplyr::pull(TF) #only th TFs

# build matrix: TF vs cluster
top_acts_mat <- df_tf %>%
  dplyr::filter(TF %in% top_tfs) %>% #filter for only the top tfs selected before
  tidyr::pivot_wider(
    id_cols = cluster,
    names_from = TF,
    values_from = mean_activity
  ) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix() %>%
  t() #tanspose

# check range
range(top_acts_mat, na.rm = TRUE)
#[1] -3.003488  3.406092

# color scale, dynamic, based on data
palette_length <- 100
my_color <- colorRampPalette(c("blue", "white", "red"))(palette_length)

# heatmap breaks set
my_breaks <- c(
  seq(-3, 0, length.out = ceiling(palette_length / 2) + 1),
  seq(0.05, 3, length.out = floor(palette_length / 2))
)

####
###
# Heatmap

# file
heatmap_base <- make_path(dir = dir_plots, filename = paste0("TF_activity_heatmap_manualAnnot_top", n_tfs))

## PNG
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = TRUE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 90, # rotate
  filename = paste0(heatmap_base, ".png"),
  width = 6,
  height = 0.15 * n_tfs
)


## Pdf
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = TRUE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 90, # angle annot
  filename = paste0(heatmap_base, ".pdf"),
  width = 6,
  height = 0.15 * n_tfs
)

```

### Return Idents
```{r}
Idents(obj) <- "seurat_clusters"
table(Idents(obj))
```


## Locked heatmap

Locked order of clusters

```{r}
# idents are seurat clusters
Idents(obj) <- "seurat_clusters"
table(Idents(obj))

# how many top TFs to get
n_tfs <- 100

# get ULM assay
DefaultAssay(obj) <- "tfsulm"

# get TF activity matrx
# Note: tfsulm@data contains scaled ULM activities
# (scale.data was intentionally copied into data earlier)
mat <- GetAssayData(obj, assay = "tfsulm", slot = "data")

## df pivot
# long: cell -> cluster -> TF
df_tf <- t(as.matrix(mat)) %>% #transpose so cells are rows 
  as.data.frame() %>%
  dplyr::mutate(cluster = Idents(obj)) %>%
  tidyr::pivot_longer( #long df
    cols = -cluster,
    names_to = "TF",
    values_to = "activity"
  ) %>%
  dplyr::group_by(cluster, TF) %>%
  dplyr::summarise(mean_activity = mean(activity), .groups = "drop")

# get top TFs by variability across clusters
top_tfs <- df_tf %>%
  dplyr::group_by(TF) %>%
  dplyr::summarise(sd_across_clusters = sd(mean_activity), .groups = "drop") %>%
  dplyr::arrange(desc(abs(sd_across_clusters))) %>%
  dplyr::slice_head(n = n_tfs) %>%
  dplyr::pull(TF) #only th TFs

# build matrix: TF vs cluster
top_acts_mat <- df_tf %>%
  dplyr::filter(TF %in% top_tfs) %>% #filter for only the top tfs selected before
  tidyr::pivot_wider(
    id_cols = cluster,
    names_from = TF,
    values_from = mean_activity
  ) %>%
  tibble::column_to_rownames("cluster") %>%
  as.matrix() %>%
  t() #tanspose

# check range
range(top_acts_mat, na.rm = TRUE)
#[1] -3.003488  3.406092

# color scale, dynamic, based on data
palette_length <- 100
my_color <- colorRampPalette(c("blue", "white", "red"))(palette_length)

# heatmap breaks set
my_breaks <- c(
  seq(-3, 0, length.out = ceiling(palette_length / 2) + 1),
  seq(0.05, 3, length.out = floor(palette_length / 2))
)

####
###
# Heatmap

# file
heatmap_base <- make_path(dir = dir_plots, filename = paste0("TF_activity_heatmap_SeuratClusters_fixedOrder_top", n_tfs))

## PNG
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = FALSE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 0, # dont rotate cluster labels here
  filename = paste0(heatmap_base, ".png"),
  width = 6,
  height = 0.15 * n_tfs
)

## PNG
pheatmap(
  mat = top_acts_mat,
  border_color  = NA,
  color = my_color,
  breaks = my_breaks,
  cluster_rows = TRUE, # hierarchically clustered TFs
  cluster_cols = FALSE, #default is true! that's why cluster doesnt follow num order
  fontsize_row = 10, #tf name size
  fontsize_col = 10, #cluster name size
  angle_col = 0, # dont rotate cluster labels here
  filename = paste0(heatmap_base, ".pdf"),
  width = 6,
  height = 0.15 * n_tfs
)


```


# Save RDS
```{r}
obj_file <- file.path(
  dir_obj,
  paste0("07_", TISSUE, "_IntestinalSeurat_TFactivity.rds")
)

saveRDS(obj, obj_file)
```

