---
title: "HDAC1 and HDAC2 Intestinal Immunity - CD4 T cell reclustering"
author: "Aruana Hansel Fröse, Moritz Madern, Phuc Tran"
date: '`r format(Sys.Date())`'
output: 
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    theme: cerulean
---

# Reclustering of CD4+ T Cells

Here we will retrieve only the CD4+ T cells from the specific tissue and recluster this population.

For specific intestinal tissues we will get:

* SI-IEL: CD4+, CD4+ CD8aa, Cluster 3


The script `06_Intestinal_scRNAseq_ModuleScores_MarkerGenes.Rmd` produces the RDS object that is necessary for this downstream analysis.

Here it is possible to select one out of the three tissues and run the downstream analysis. Refer to the *parameters*, *dirs* and *tissue* section.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries  
```{r, message=FALSE, warnings=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(RColorBrewer)
library(knitr)
library(Seurat)
library(SeuratData)
library(readxl)
library(openxlsx)
library(ggrepel)

library(org.Mm.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(clusterProfiler)
library(rrvgo)
library(stats)
library(enrichplot)

```

# Setup


## TISSUE
```{r}
TISSUE <- "SI-IEL"

stopifnot(TISSUE %in% c("SI-IEL", "SI-LPL", "colon"))
```

## Object source

By default, manuscript figures are reproduced using the frozen Seurat object used for the final analysis. Set `use_frozen_object = FALSE` to use the object generated by the preceding pipeline scripts.

*Minor differences in stochastic dimensionality-reduction results may occur when the full workflow is rerun from raw data.*

```{r}
use_frozen_object <- TRUE
```

## Directories
```{r}
dir_input <- here("Data", "Processed", "06_ModuleScores", TISSUE)
dir_frozen <- here("Data", "Processed", "00_Frozen")

dir_metadata <- here("Data", "Processed", "Additional_metadata")

dir_res <- here("Results", "08_CD4_reclustering", TISSUE) #parent dir
dir_obj <- here("Data", "Processed", "08_CD4_reclustering", TISSUE) # objects

dir_plots <- file.path(dir_res, "Plots")
dir_tables <- file.path(dir_res, "Tables")

dirs <- c(dir_obj, dir_plots, dir_tables)

lapply(dirs, dir.create, recursive = TRUE, showWarnings = FALSE)
```

# Load data
```{r}
frozen_obj_file <- file.path(
  dir_frozen,
  "06_SI-IEL_IntestinalSeurat_integrated_ModuleScoresManualAnnot_FROZEN.rds"
)

obj_file <- file.path(
  dir_input,
  paste0("06_", TISSUE, "_IntestinalSeurat_integrated_ModuleScoresManualAnnot.rds")
)

if (use_frozen_object) {
  stopifnot(TISSUE == "SI-IEL")

  obj <- readRDS(frozen_obj_file)
  message("Using frozen Seurat object for exact manuscript figure reproduction.")
  
} else {
  obj <- readRDS(obj_file)
  message("Using Seurat object generated by the reproducible analysis pipeline.")

  }
```

```{r}
obj
# An object of class Seurat 
# 39647 features across 6071 samples within 5 assays 
# Active assay: SCT (15709 features, 0 variable features)
#  3 layers present: counts, data, scale.data
#  4 other assays present: RNA, HTO, CITE, integrated
#  2 dimensional reductions calculated: pca, umap
```

## CELL prefix is [CD4]
This cell prefix will be used for names or files saved here, such as CD4 subset or CD8 subset, etc.

```{r}
CELL <- "CD4"
```


## Helper functions
Helper function for creating new paths for new output

```{r}
## Make path
make_path <- function(dir, filename) file.path(dir, paste0(TISSUE,"_", CELL, "_", filename))

# save all extensions for the plots 
save_plot <- function(plot, dir, name, w = 10, h = 6) {
  base <- file.path(dir, paste0(TISSUE, "_", CELL, "_", name))
  ggsave(paste0(base, ".pdf"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".png"), plot = plot, width = w, height = h)
  ggsave(paste0(base, ".svg"), plot = plot, width = w, height = h)
}
```


## Tissue dictionary
```{r}
# map the tissue to the specific HTO classes considering the genotype
tissue_dict <- list(
  "SI-IEL" = c("WT-SI-IEL",  "KO-SI-IEL"),
  "SI-LPL" = c("WT-SI-LPL",  "KO-SI-LPL"),
  "colon"  = c("WT-colon",   "KO-colon")
)

tissue_dict

```

## Subset parameters
```{r}
# get the specific WT/KO from the selected tissue
subselection_HTOclasses <- tissue_dict[[TISSUE]]
# check
message("These are the genotypes measured in this tissue: ")
print(subselection_HTOclasses)

# generate groups for Differential Analysis pairwise comparisons in correct order (KO first, WT second)
DEgroups_HTOclasses <- list(
  list(
    c(subselection_HTOclasses[2]),   # KO first
    c(subselection_HTOclasses[1])    # WT second
  )
)
message("These are the pairwise comparison for further DE analysis:")
print(DEgroups_HTOclasses)
```

## Cluster parameters
```{r}
set.seed(42)

## Specify the number of PCA dimensions to be used for 1) dimension reduction via UMAP, and 2) clustering
numberDimsPCA_for_UMAP = 8

## Specify the minimum distance (first main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
minDist_for_UMAP = 0.3


## Specify the spread (second main parameter used for UMAP calculation) to be used for dimension reduction via UMAP
spread_for_UMAP = 1


## Specify number of nearest neighbours for dimension reduction via UMAP
n_neighbours_for_UMAP = 7


## Specify local connectivity for dimension reduction via UMAP
local_connectivity_for_UMAP = 20


## Specify negative sample rate for dimension reduction via UMAP
negative_sample_rate_for_UMAP = 20


## Specify the resolution to be used for clustering. A lower resolution will produce a lower number of clusters
res_for_clustering = 0.5
```


## Module Scores
```{r}
# Module Scores genes
# Load your CSV file (with cluster_name in column 1 and gene_name in column 2)
fileannot <- file.path(dir_metadata,"Annotation_final_list_for_module_score.csv")

geneSet_df <- read.csv(fileannot, header = FALSE, stringsAsFactors = FALSE)

```

## Genes of interest
```{r}
genes_of_interest <- c("Sell","Il7r","Il6ra","Ccr7","Runx1","Klf2","Icos","Cd28",
                       "Tcf7","Cd7","Foxp1","Ikzf1","Gzma","Runx3","Tbx21","Fgl2",
                       "Ccr5","Ccl4","Mki67","Ccl5","Tgfb1","Ikzf2","Ccl3",
                       "Gzmb","Klrb1b","Cd53","Hdac1","Cd8a","Cd8b1","Ccr5","Nkg7",
                       "Cxcr5","Ctla4","Ikzf3","Cd200r2","Ccr9","Crtam", "Lef1", "Cd5",
                       "Cd6", "Btla", "Cd160", "Cd274", "Socs1", "Pycard", "Hspb1")

```

## Biomart
```{r}
biomart_tab <- read.csv(file.path(dir_metadata, "Ensembl_115_biomart_mm10_biotype.csv"))
```

## Human gene list

Obtained from the reference PMC5931334

```{r message=FALSE, warning=FALSE}
gene_list <- readLines(file.path(dir_metadata, "human_CD4_CTL_gene_set_from_PMC5931334.txt"))
```


***

# Recluster CD4

Recluster CD4 T cell, including CD4+CD8aa and cluster 3 from SI-IEL

## CD4 level
Level of CD4 expression to choose

```{r}

DefaultAssay(obj) <- "CITE"

VlnPlot(obj, features = CELL, group.by = "seurat_clusters")

FeaturePlot(obj, features = CELL, label = TRUE)

# CD4 > 1
```

## Subset [CD4]
```{r}
# Step 1: Set Default Assay to CITE
DefaultAssay(obj) <- "CITE"

# Step 2: Set identity to clusters
obj <- SetIdent(object = obj, 
                value = obj$seurat_clusters)

# Step 3: Subset cells
# keep all cells that are in cluster 3, and cells that are CD4 (from CITEseq assay) signal is above 1, from clusters 0 to 12.
obj.cell <- subset(x = obj,
                   subset = (
                     (seurat_clusters %in% c("3")) | # all cells in cluster 3 regardless of CD4 citeseq
                       ((CD4 > 1) & seurat_clusters %in% c("0","1","2","4","5","6","7", "8","9","10","11","12")) 
                   )
)

# check
obj.cell
# An object of class Seurat 
# 39647 features across 2305 samples within 5 assays 
# Active assay: CITE (3 features, 0 variable features)
#  2 layers present: counts, data
#  4 other assays present: RNA, HTO, SCT, integrated
#  2 dimensional reductions calculated: pca, umap

# Step 4: Switch to integrated assay
DefaultAssay(obj.cell) <- "integrated"

# Step 5: PCA on variable features
obj.cell <- RunPCA(
  obj.cell,
  assay = "integrated",
  features = VariableFeatures(obj.cell, assay = "integrated")
)


DimPlot(obj.cell, reduction = "umap", label = TRUE)

```

## Overview
Here is an overview of the data after subselection based on CD4 clusters:

```{r}
# summary numbers
obj.cell
# An object of class Seurat 
# 39647 features across 2305 samples within 5 assays 
# Active assay: integrated (3000 features, 3000 variable features)
#  2 layers present: data, scale.data
#  4 other assays present: RNA, HTO, CITE, SCT
#  2 dimensional reductions calculated: pca, umap

## Print Experiments
writeLines("Experiments:")
obj.cell$Experiment %>% unique %>% sort()

writeLines("Cells per experiment:")
obj.cell$Experiment %>% table() %>% print()


## Print Tissues
writeLines("Tissues:")
obj.cell@meta.data$tissue %>% unique() %>% as.character() %>% sort()


## Print HTO classes
writeLines("HTO classes:")
obj.cell@meta.data$HTO_classification %>% unique() %>% as.character() %>% sort()

writeLines("Cells per HTO class:")
obj.cell@meta.data$HTO_classification %>% table() %>% print()


## Redefine HTO classification factor
obj.cell$HTO_classification <- factor(obj.cell$HTO_classification, levels=subselection_HTOclasses)


## Print CITE-seq quantified proteins
writeLines("CITE-seq targets:")
obj.cell@assays$CITE$counts %>% rownames()


```

# Reintegration of subset

Note that the integration step of the two experiments needs to be performed again as other genes might explain most of the variation in the new subset of cells. In other words, the set of most-variable genes might differ from the original set of genes, on which basis PCA and UMAP are calculated. Here are the results after the new integration and subsequent recalculation of UMAP dimension reduction and PCA:

```{r}
## Delete previous "SCT" and "integrated" assays
DefaultAssay(obj.cell) <- "RNA"

obj.cell[["SCT"]] <- NULL
obj.cell[["integrated"]] <- NULL


## Subset for each individual experiment and perform SC transform
Experiment1 <- subset(obj.cell, subset = Experiment == "Experiment1")
Experiment2 <- subset(obj.cell, subset = Experiment == "Experiment2")

Experiment1 <- SCTransform(Experiment1)
Experiment2 <- SCTransform(Experiment2)

obj.cell_list <- list(Experiment1, Experiment2)

```

```{r message=FALSE, warning=FALSE, include = FALSE}
# clean
rm(list=c("Experiment1", "Experiment2"))
gc()
```

## SCT integration

```{r message=FALSE, warning=FALSE}
##  Perform Integration
IntegFeatures <- SelectIntegrationFeatures(object.list = obj.cell_list,
                                           nfeatures=3000)

obj.cell_list <- PrepSCTIntegration(object.list = obj.cell_list, 
                                      anchor.features = IntegFeatures)

IntegAnchors <- FindIntegrationAnchors(object.list = obj.cell_list,
                                       normalization.method = "SCT",
                                       anchor.features = IntegFeatures)
## new integrated seurat obj
int.obj.cell <- IntegrateData(anchorset = IntegAnchors,
                                 normalization.method = "SCT")
```

## Check

```{r}
# after integration
DefaultAssay(int.obj.cell)

Assays(int.obj.cell)

nrow(GetAssayData(int.obj.cell, assay = "SCT"))

nrow(GetAssayData(int.obj.cell, assay = "integrated"))

length(IntegFeatures)

all(IntegFeatures %in% rownames(int.obj.cell[["integrated"]]))

head(VariableFeatures(int.obj.cell[["SCT"]]))

head(VariableFeatures(int.obj.cell[["integrated"]]))


```

## Compute missing SCT variableFEatures
```{r}
VariableFeatures(int.obj.cell[["SCT"]]) <- IntegFeatures

head(VariableFeatures(int.obj.cell[["SCT"]]))

```



# Plots overview

## Before integration 
assay is SCT

```{r}
## PCA/UMAP before & after integration for one tissue

# PCA before (assay = "SCT")
DefaultAssay(int.obj.cell) <- "SCT"

# RunPCA
int.obj.cell <- RunPCA(
  int.obj.cell,
  assay   = "SCT",
  features = VariableFeatures(int.obj.cell[["SCT"]]),
  reduction.name = "pca.unintegrated"   # save it to obj
)

# DimPlot
pca.bf <- DimPlot(
  int.obj.cell,
  reduction = "pca.unintegrated",
  group.by  = "Experiment"
) + ggtitle(paste("PCA before integration -", TISSUE))



# UMAP before
# RunUmap
int.obj.cell <- RunUMAP(
  int.obj.cell,
  reduction            = "pca.unintegrated", # which one to use
  reduction.name       = "umap.unintegrated", # which name to store it
  dims                 = 1:numberDimsPCA_for_UMAP, # parameter set up in Parameters
  min.dist             = minDist_for_UMAP, # also set
  spread               = spread_for_UMAP, # set before
  n.neighbors          = n_neighbours_for_UMAP, # set 
  local.connectivity   = local_connectivity_for_UMAP, # set
  negative.sample.rate = negative_sample_rate_for_UMAP # set
)


# DimPlot
umap.bf <- DimPlot(
  int.obj.cell,
  reduction = "umap.unintegrated",
  group.by  = "Experiment"
) + ggtitle(paste("UMAP before integration -", TISSUE))


```

## After integration
assay is integrated

```{r}
####### AFTER integration
# PCA after (assay = "integrated")
DefaultAssay(int.obj.cell) <- "integrated"

# RunPCA
int.obj.cell <- RunPCA(
  int.obj.cell,
  assay    = "integrated",
  reduction.name = "pca.integrated",   # save it to obj
  features = VariableFeatures(int.obj.cell[["integrated"]])
)

pca.af <- DimPlot(
  int.obj.cell,
  reduction = "pca.integrated",
  group.by  = "Experiment"
) + ggtitle(paste("PCA after integration -", TISSUE))

ggsave(
  filename = file.path(dir_plots, paste0(TISSUE, "_", CELL, "_PCA_after_integration.png")),
  plot = pca.af, width = 10)

# UMAP AFTER
# Run UMAP
int.obj.cell <- RunUMAP(
  int.obj.cell,
  reduction            = "pca.integrated", # which one to use
  reduction.name       = "umap.integrated", # which name to save
  dims                 = 1:numberDimsPCA_for_UMAP,
  min.dist             = minDist_for_UMAP,
  spread               = spread_for_UMAP,
  n.neighbors          = n_neighbours_for_UMAP,
  local.connectivity   = local_connectivity_for_UMAP,
  negative.sample.rate = negative_sample_rate_for_UMAP
)

# DImplot
umap.af <- DimPlot(
  int.obj.cell,
  reduction = "umap.integrated",
  group.by  = "Experiment"
) + ggtitle(paste("UMAP after integration -", TISSUE))

ggsave(
  filename = file.path(dir_plots, paste0(TISSUE, "_", CELL, "_UMAP_after_integration.png")),
  plot = umap.af, width = 10)

```

## All plots
```{r}
# overview.pcas <- pca.bf | pca.af
# 
# ggsave(filename = file.path(dir_plots,
#                             paste0(TISSUE, "_", CELL, "_PCAs_before_after_overview.png")),
#        plot = overview.pcas, width = 20)
# 
# 
# overview.umaps <- umap.bf | umap.af
# 
# ggsave(filename = file.path(dir_plots, paste0(TISSUE, "_", CELL, "_UMAPs_before_after_overview.png")),
#        plot = overview.umaps,  width = 20)

pca.bf
pca.af

umap.bf
umap.af


```


## ElbowPlot
```{r}
writeLines(paste("Elbow plot to determine appropriate number of PCA dimensions used for UMAP (", numberDimsPCA_for_UMAP, ")"))


ElbowPlot(int.obj.cell, reduction="pca.integrated", ndims=50) + ggtitle(paste("Elbow plot integrated", TISSUE))

```

## JoinLayers

```{r}
int.obj.cell[["RNA"]] <- JoinLayers(int.obj.cell[["RNA"]])

int.obj.cell
```

## Save RDS 
```{r}
obj_file <- file.path(dir_obj, paste0("08_", TISSUE, "_", CELL, "_subset_afterIntegration.rds"))

if (use_frozen_object) {
  stopifnot(TISSUE == "SI-IEL")

  frozen_obj_file <- file.path(dir_frozen, "08_SI-IEL_CD4_subset_afterIntegration_FROZEN.rds")
  
  int.obj.cell <- LoadSeuratRds(frozen_obj_file)
  message("Using frozen CD4 reintegrated object for exact manuscript figure reproduction.")

} else {
  saveRDS(int.obj.cell, obj_file)
  message("Using CD4 reintegrated object generated by the reproducible analysis pipeline.")
}
```


# Clustering

```{r}
# Change default assay to integrated
DefaultAssay(int.obj.cell) <- "integrated"

# Calculate PCA and perform clustering
int.obj.cell <- RunPCA(
  int.obj.cell, 
  assay = "integrated", 
  reduction.name = "pca.integrated",   # save it to obj
  features = VariableFeatures(int.obj.cell, assay = "integrated")
)

int.obj.cell <- FindNeighbors(int.obj.cell, dims = 1:numberDimsPCA_for_UMAP)

int.obj.cell <- FindClusters(int.obj.cell, resolution = res_for_clustering)

# Ensure genotype order (WT left, KO right)
int.obj.cell$genotype <- factor(int.obj.cell$genotype, levels = c("WT", "KO"))

# Run UMAP
int.obj.cell <- RunUMAP(
  int.obj.cell,
  reduction            = "pca.integrated", # which one to use
  reduction.name       = "umap.integrated", # which name to save
  dims = 1:numberDimsPCA_for_UMAP,
  min.dist = minDist_for_UMAP,
  spread = spread_for_UMAP,
  n.neighbors = n_neighbours_for_UMAP,
  local.connectivity = local_connectivity_for_UMAP,
  negative.sample.rate = negative_sample_rate_for_UMAP
)

# Generate plots
p1 <- DimPlot(
  int.obj.cell, 
  reduction = "umap.integrated", 
  split.by = "genotype", 
  group.by = "seurat_clusters", 
  label = TRUE
) + ggtitle(paste("UMAP after integration ", TISSUE, CELL, "subset"))

print(p1)

# Save plots as SVG
filename <- make_path(dir = dir_plots, filename = "UMAP_clusters_split")
filename

ggsave(paste0(filename, ".pdf"), plot = p1, width = 10, height = 6, device = "pdf")
ggsave(paste0(filename, ".svg"), plot = p1, width = 10, height = 6, device = "svg")
ggsave(paste0(filename, ".png"), plot = p1, width = 10, height = 6, device = "png")

```


# CITEseq

Note: The following UMAP dimension reduction was calculated on the basis of the RNA expression (count) data.


```{r}
DefaultAssay(int.obj.cell) <- "CITE"

for (i in rownames(int.obj.cell@assays$CITE$data)) {
  message("Processing: ", i)

  # Generate feature plot
  p <- FeaturePlot(
    int.obj.cell,
    features = i,
    cols = c("lightgrey", "darkblue"),
    slot = "data"
  ) + ggtitle(paste0("", i)) 
    # theme_void() +  # removes background, grid, axes
    # theme(
    #   plot.title = element_text(size = 50, hjust = 0.7, vjust = -25),
    #   legend.position = "NA"
    # )

  # Show plot in session
  print(p)

  # Save each plot
  ggsave(
    filename = make_path(dir = dir_plots, filename = paste0("FeaturePlot_CITEseq_", i, ".png")),
    plot = p,
    width = 8, height = 6
  )

  # Clean up
  rm(p)
  gc()
}

DefaultAssay(int.obj.cell) <- "integrated"
```


# Module Scores

Create Module Scores based on Custom Gene Sets for annotation (Figure S2D)

## JoinLayers

```{r warning=FALSE}
# make sure layers are joined
int.obj.cell[["RNA"]] <- JoinLayers(int.obj.cell[["RNA"]])

```


```{r}
# Set default assay
DefaultAssay(int.obj.cell) <- "RNA"

## Extract feature names
featurenames <- int.obj.cell@assays$RNA$data %>% rownames()
featurenames_integrated <- int.obj.cell@assays$integrated$data %>% rownames()


# Optional: Clean whitespace
geneSet_df$cluster_name <- trimws(geneSet_df[[1]])
geneSet_df$gene_name <- trimws(geneSet_df[[2]])

# Remove empty rows or missing genes
geneSet_df <- geneSet_df[geneSet_df$gene_name != "" & !is.na(geneSet_df$gene_name), ]

# Group genes by cluster
gene_sets <- split(geneSet_df$gene_name, geneSet_df$cluster_name)

# Loop through each gene set (per cluster)
for (cluster in names(gene_sets)) {
  
  customGeneSet <- gene_sets[[cluster]]
  
  # Filter genes to those present in the dataset
  customGeneSet <- customGeneSet[customGeneSet %in% featurenames]
  
  if (length(customGeneSet) == 0) {
    message(paste0("No matching genes found in dataset for cluster: ", cluster))
    next
  }
  
  # Print info
  message("Cluster: ", cluster)
  message("Number of genes found in data: ", length(customGeneSet))
  print(head(customGeneSet, 100))
  
  # Calculate module score
  int.obj.cell <- AddModuleScore(
    object = int.obj.cell,
    features = list(customGeneSet),
    name = cluster,
    assay = "RNA",
    slot = "data"
  )
  
  # Create FeaturePlot
  feature_to_plot <- paste0(cluster, "1")
  
  gg <- FeaturePlot(
    object = int.obj.cell,
    features = feature_to_plot
  ) + 
    scale_colour_gradientn(colours = rev(brewer.pal(n = 11, name = "RdBu"))) +
    ggtitle(cluster)
    # theme(
    #   plot.title = element_text(size = 35,
    #                             face = "plain",
    #                             hjust = 0.5,
    #                             vjust = -35), # centered and placed below
    # legend.position = "none" , # removes the legend
    # axis.title = element_blank()
  # )
  
 
  # Save plot
  filename <- make_path(dir = dir_plots, filename = paste0("FeaturePlot_", cluster))
  filename
  
  # Save the plot
  ggsave(paste0(filename, ".png"), plot = gg, width = 6, height = 5, dpi = 500)
  ggsave(paste0(filename, ".svg"), plot = gg, width = 6, height = 5, dpi = 500)

  print(gg)
  
  rm(gg)
  gc()
}

```



# Marker Genes for Clusters 

Using Seurat’s FindMarkers function, we can test for gene-wise differential expression between a specific cell cluster and its respective complement (i.e. all cells from all other clusters). The DE-testing results are filtered for positive fold changes (i.e. gene is up in the respective cluster) and the expression pattern of the top 30 genes per test are illustrated by heatmap. The full table results for each cluster, containing all significant genes with positive fold changes per cluster, are stored in the "Results" directory.

```{r}
## Set identification and extract loop variables (the clusters)
Idents(int.obj.cell) <- "seurat_clusters"
loop_vars <- Idents(int.obj.cell) %>% levels() 

# list to save markers per cluster
top_markers_list <- list()
complete_markers_list <- list()

#### 
## FindMarkers
DefaultAssay(int.obj.cell) <- "SCT"
# find markers that define tissue types using FindMarkers function on the SCT assay
int.obj.cell <- PrepSCTFindMarkers(int.obj.cell)

## Perform loop (differential analysis per cluster vs rest)
for (i in loop_vars){
  
  # print
  message(paste("FindMarkers of Cluster",i))
  
  # get Markers from this cluster only 
  resultFindMarkers <- FindMarkers(int.obj.cell, assay = "SCT", ident.1 = i)
  
  # subselect for positive Fold change
  # if there are any
  if (nrow(resultFindMarkers) > 0) {
    resultFindMarkers <- resultFindMarkers %>% filter(avg_log2FC > 0 &  p_val_adj < 0.05)
  }
  
  # gene names as new column
  GeneNames <- resultFindMarkers %>% rownames()
  resultFindMarkers <- cbind(GeneNames, resultFindMarkers)

   # add cluster name (same for all rows)
  resultFindMarkers$Cluster <- i

  # select top results (for subsequent heatmap plotting)
  # if there are any significant
  if (nrow(resultFindMarkers) > 0) {
    resultFindMarkers_top30 <- resultFindMarkers %>%
        slice_head(n = 30) #get top 30 even when n < 30 and n = 0.
  
    # save markers and top30 to list
    complete_markers_list[[i]] <- resultFindMarkers
    
    top_markers_list[[i]] <- resultFindMarkers_top30
    
  }
  
# save result as table
  filename_table <- make_path(dir = dir_tables, filename = paste0("MarkerGenes_padj005_Cluster_", i,".txt"))

  write.table(x = resultFindMarkers, 
              file = filename_table,
              col.names =TRUE,
              row.names = FALSE, 
              quote=FALSE, 
              sep="\t")
  
}

```

```{r}
gc()
```



# Diff exp KO vs WT

The groups for differential expression testing by pairwise group comparison KO vs WT are specified by the parameter "`DEgroups_HTOclasses`".

Significant genes are defined as: **adj.p-value < 0.05**.

## Prep Setup
```{r}
## Set identification and adjust default assay
Idents(int.obj.cell) <- "HTO_classification"
DefaultAssay(int.obj.cell) <- "SCT"

# Prepare SCT assay once for DE
int.obj.cell <- PrepSCTFindMarkers(int.obj.cell)

```

## Mouse orthologs

Define gene lists for which enrichment testing across our GSEA ranking should be conducted

```{r}
# MSigDB gene sets for GSEA
df_msig_hallmark <- msigdbr(
  species  = "Mus musculus",
  category = "H"
) %>%
  dplyr::select(gs_name, gene_symbol) %>%
  as.data.frame()

df_msig_go <- msigdbr(
  species  = "Mus musculus",
  category = "C5" #GO BP
) %>%
  dplyr::filter(gs_subcat %in% "GO:BP") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  as.data.frame()

df_geneSets <- rbind(df_msig_hallmark, df_msig_go)

```

## KO vs WT DEA loop
```{r message=FALSE, warning=FALSE}
if (!is.null(DEgroups_HTOclasses)) {
  for (i in seq_along(DEgroups_HTOclasses)) {
    
    # extract groups
    group1 <- DEgroups_HTOclasses[[i]][[1]]  # KO
    group2 <- DEgroups_HTOclasses[[i]][[2]]  # WT
    
    group1_collapsed <- paste(group1, collapse = " ")
    group2_collapsed <- paste(group2, collapse = " ")
    PairwiseGroups_title <- paste0(group1_collapsed, "_vs_", group2_collapsed)
    
    writeLines(paste0("Pairwise Group Comparison ", i, ":"))
    writeLines(paste0("Group1: ", group1_collapsed))
    writeLines("versus")
    writeLines(paste0("Group2: ", group2_collapsed))
    
    # FindMarkers
    # DE test: group1 vs group2
    resultFindMarkers <- FindMarkers(
      int.obj.cell,
      assay   = "SCT",
      ident.1 = group1,
      ident.2 = group2
    )
    geneNames <- rownames(resultFindMarkers)
    resultFindMarkers <- cbind(geneNames, resultFindMarkers)
    
    ## Prepare volcano plot data
    resultFindMarkers$neglog_p_val_adj <- -log10(resultFindMarkers$p_val_adj)
    resultFindMarkers$neglog_p_val_adj[resultFindMarkers$neglog_p_val_adj > 300] <- 300
    # gene labels only if significant, otherwise empty
    resultFindMarkers$DElabel <- ifelse(resultFindMarkers$p_val_adj< 0.05, 
                                        yes = resultFindMarkers$geneNames,
                                        no = "")
    
    # significance
    resultFindMarkers$significant <- ifelse(resultFindMarkers$p_val_adj < 0.05, "+", "-")
    # colors
    resultFindMarkers$colors_ggplot <- "ColGrey"
    resultFindMarkers$colors_ggplot[resultFindMarkers$significant == "+"] <- "ColRed"
    resultFindMarkers$colors_ggplot[abs(resultFindMarkers$avg_log2FC) < 0.25 & resultFindMarkers$p_val_adj < 0.05] <- "ColBlue"
    
    resultFindMarkers$colors_ggplot <- factor(resultFindMarkers$colors_ggplot,levels = c("ColGrey", "ColBlue", "ColRed"))
    
    # symmetric x-limits based on largest absolute FC
    xbound <- max(abs(resultFindMarkers$avg_log2FC), na.rm = TRUE)
    
    # highlight genes of interest that are significant in red
    highlight_df <- resultFindMarkers %>%
      dplyr::filter(
        geneNames %in% genes_of_interest,
        colors_ggplot == "ColRed"
      )
    
    #### 
    # Volcano plot with diagonal bars for highlighted red custom genes
    gg <- ggplot(data = resultFindMarkers,
                 mapping = aes(x = avg_log2FC,
                               y = neglog_p_val_adj)
    ) +
      geom_point(aes(color = colors_ggplot), alpha = 0.9, size  = 1.8
                 ) +
     scale_color_manual(
  values = c("ColGrey" = "grey",
             "ColBlue" = "blue",
             "ColRed"  = "red"),
  breaks = c("ColGrey", "ColBlue", "ColRed"),
  labels = c("ns",
             "padj < 0.05",
             "padj < 0.05 and |log2FC| ≥ 0.25"),
  name = ""

  ) +
      geom_text_repel( #repel labels dinamically
        data = highlight_df,
        aes(x = avg_log2FC,
            y = neglog_p_val_adj,
            label = geneNames),
        size = 4,
        fontface = "italic",
        box.padding = 0.4,
        point.padding = 0.3,
        max.overlaps = Inf,
        min.segment.length = 0.2,   # keeps small connecting segments
        segment.color = "black",
        segment.size = 0.4
      ) +
      geom_hline(  # Add threshold line
        yintercept = -log10(0.05),
        lty = 2,
        color = "black"
      ) +
      geom_vline(  # Add threshold line
        xintercept = c(-0.25, 0.25),
        lty = 2,
        color = "black"
      ) +
      ggtitle(paste(PairwiseGroups_title, "padj 0.05")) +
      xlim(-xbound, xbound) +
      xlab("Average log2 fold change") +
      ylab("-log10 adjusted p-value") +
      theme_bw(base_size = 12) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
      )
    
    print(gg)
    
    # unique volcano filename per comparison
    volcano_file <- make_path(dir = dir_plots,
                              filename = paste0("VolcanoPlot_",
                                                PairwiseGroups_title, "_padj005"))

    ggsave(paste0(volcano_file, ".pdf"), plot = gg, width = 10, height = 6)
    ggsave(paste0(volcano_file, ".png"), plot = gg, width = 10, height = 6)
    ggsave(paste0(volcano_file, ".svg"), plot = gg, width = 10, height = 6)

    ################
    ## Results table
    
    # Add protein coding info to the res from biomart
    resultFindMarkers <- merge(
      x    = resultFindMarkers,
      y    = biomart_tab,
      all.x = TRUE,
      by.x  = "geneNames",
      by.y  = "external_gene_name"
    )
    
    # Filter significant genes 
     resultFindMarkers_sig <- resultFindMarkers %>%
      dplyr::filter(p_val_adj < 0.01)
     
    filename_table <- make_path(dir_tables,
                                   filename = paste0("MarkerGenes_padj001_",
                                   PairwiseGroups_title,
                                   ".txt")
    )
    
    write.table(x = resultFindMarkers_sig,
                file = filename_table,
                col.names = TRUE, row.names = FALSE,quote= FALSE,sep= "\t")
    
    ### padj 005
    resultFindMarkers_sig005 <- resultFindMarkers %>%
      dplyr::filter(p_val_adj < 0.05)
     
    filename_table005 <- make_path(dir_tables,
                                   filename = paste0("MarkerGenes_padj005_",
                                   PairwiseGroups_title,
                                   ".txt")
    )
    
    write.table(x = resultFindMarkers_sig005,
                file = filename_table005,
                col.names = TRUE, row.names = FALSE,quote= FALSE,sep= "\t")
    
    #########
    ## GSEA
    # calculate ranking for GSEA:
    # ranking all genes, filtered to protein_coding)
    ranked_query_list <- setNames(
      resultFindMarkers$avg_log2FC,
      nm = resultFindMarkers$geneNames
    )
    
    # protein coding & unique symbols
    # pull geneNames
    proteinCoding <- resultFindMarkers %>%
      dplyr::filter(gene_biotype == "protein_coding") %>%
      dplyr::distinct(geneNames) %>%
      dplyr::pull(geneNames)
    
    # no duplicates and NAs
    ranked_query_list <- ranked_query_list[proteinCoding]
    ranked_query_list <- ranked_query_list[!is.na(ranked_query_list)]
    ranked_query_list <- ranked_query_list[!duplicated(names(ranked_query_list))]
    # sort rank
    ranked_query_list <- sort(ranked_query_list, decreasing = TRUE)
    
    ### GSEA
    # skip if too few genes
    if (length(ranked_query_list) < 20) {
      message("Skipping GSEA for ", PairwiseGroups_title, " – less than 20 genes")
      next
    }
    # perform GSEA    
    gsea_res <- GSEA(
      geneList = ranked_query_list,
      TERM2GENE = df_geneSets
    )
    df_gsea_res <- gsea_res@result
    
    filename_table <- make_path(dir_tables,
                                   filename = paste0(
                                     "MarkerGenes_GSEA_",
                                     PairwiseGroups_title,
                                     ".txt")
    )
    
    write.table(x = df_gsea_res,
                file = filename_table,
                col.names = TRUE,
                row.names = FALSE,
                quote = FALSE,
                sep = "\t"
    )
  }
}

```


## How many genes up in KO and WT

```{r}
up_down_counts <- resultFindMarkers %>%
  filter(colors_ggplot == "ColRed") %>%
  summarise(
    up_in_WT = sum(avg_log2FC <  0),
    up_in_KO   = sum(avg_log2FC > 0)
  )

message("ColRed: padj < 0.05 and |log2FC| ≥ 0.25")
print(up_down_counts)

```

```{r}
resultFindMarkers_red <- resultFindMarkers %>%
  dplyr::filter(colors_ggplot == "ColRed")

table(resultFindMarkers_red$gene_biotype)


up_down_counts_pc <- resultFindMarkers_red %>%
  dplyr::filter(gene_biotype == "protein_coding") %>%
  dplyr::summarise(
    up_in_WT = sum(avg_log2FC < 0),
    up_in_KO = sum(avg_log2FC > 0)

  )

up_down_counts_pc

```

## Padj differences
```{r}
reds <- resultFindMarkers %>% filter(colors_ggplot == "ColRed")

reds_005 <- reds %>% summarise(
  up_in_WT = sum(avg_log2FC < 0),
  up_in_KO = sum(avg_log2FC > 0)
)

reds_001 <- reds %>% filter(p_val_adj < 0.01) %>% summarise(
  up_in_WT = sum(avg_log2FC < 0),
  up_in_KO = sum(avg_log2FC > 0)
)

print("padj 005")
print(reds_005)

print("padj 001")
print(reds_001)

```

## VolcanoPlot padj 0.01

```{r}
if (!is.null(DEgroups_HTOclasses)) {
  for (i in seq_along(DEgroups_HTOclasses)) {
    
    # extract groups
    group1 <- DEgroups_HTOclasses[[i]][[1]]  # KO
    group2 <- DEgroups_HTOclasses[[i]][[2]]  # WT
    
    group1_collapsed <- paste(group1, collapse = " ")
    group2_collapsed <- paste(group2, collapse = " ")
    PairwiseGroups_title <- paste0(group1_collapsed, "_vs_", group2_collapsed)
    
    writeLines(paste0("Pairwise Group Comparison ", i, ":"))
    writeLines(paste0("Group1: ", group1_collapsed))
    writeLines("versus")
    writeLines(paste0("Group2: ", group2_collapsed))
    
    # FindMarkers
    # DE test: group1 vs group2
    resultFindMarkers <- FindMarkers(
      int.obj.cell,
      assay   = "SCT",
      ident.1 = group1,
      ident.2 = group2
    )
    geneNames <- rownames(resultFindMarkers)
    resultFindMarkers <- cbind(geneNames, resultFindMarkers)
    
    ## Prepare volcano plot data
    resultFindMarkers$neglog_p_val_adj <- -log10(resultFindMarkers$p_val_adj)
    resultFindMarkers$neglog_p_val_adj[resultFindMarkers$neglog_p_val_adj > 300] <- 300
    # gene labels only if significant, otherwise empty
    resultFindMarkers$DElabel <- ifelse(resultFindMarkers$p_val_adj< 0.01, ########PADJ OF 0.01  
                                        yes = resultFindMarkers$geneNames,
                                        no = "")
    
    # significance
    resultFindMarkers$significant <- ifelse(resultFindMarkers$p_val_adj < 0.01, "+", "-") ###### PADJ 0.01
    # colors
    resultFindMarkers$colors_ggplot <- "ColGrey"
    resultFindMarkers$colors_ggplot[resultFindMarkers$significant == "+"] <- "ColRed"
    resultFindMarkers$colors_ggplot[abs(resultFindMarkers$avg_log2FC) < 0.25 & resultFindMarkers$p_val_adj < 0.01] <- "ColBlue" ######## PADJ 0.01
    
    resultFindMarkers$colors_ggplot <- factor(resultFindMarkers$colors_ggplot,levels = c("ColGrey", "ColBlue", "ColRed"))
    
    # symmetric x-limits based on largest absolute FC
    xbound <- max(abs(resultFindMarkers$avg_log2FC), na.rm = TRUE)
    
    # highlight genes of interest that are significant in red
    highlight_df <- resultFindMarkers %>%
      dplyr::filter(
        geneNames %in% genes_of_interest,
        colors_ggplot == "ColRed"
      )
    
    #### 
    # Volcano plot with diagonal bars for highlighted red custom genes
    gg <- ggplot(data = resultFindMarkers,
                 mapping = aes(x = avg_log2FC,
                               y = neglog_p_val_adj)
    ) +
      geom_point(aes(color = colors_ggplot), alpha = 0.9, size  = 1.8
                 ) +
     scale_color_manual(
  values = c("ColGrey" = "grey",
             "ColBlue" = "blue",
             "ColRed"  = "red"),
  breaks = c("ColGrey", "ColBlue", "ColRed"),
  labels = c("ns",
             "padj < 0.01",
             "padj < 0.01 and |log2FC| ≥ 0.25"),
  name = ""

  ) +
      geom_text_repel( #repel labels dinamically
        data = highlight_df,
        aes(x = avg_log2FC,
            y = neglog_p_val_adj,
            label = geneNames),
        size = 4,
        fontface = "italic",
        box.padding = 0.4,
        point.padding = 0.3,
        max.overlaps = Inf,
        min.segment.length = 0.2,   # keeps small connecting segments
        segment.color = "black",
        segment.size = 0.4
      ) + 
      geom_hline(  # Add threshold line
        yintercept = -log10(0.01), #### AT PADJ 0.01 ############
        lty = 2,
        color = "black"
      ) +
      geom_vline(  # Add threshold line
        xintercept = c(-0.25, 0.25),
        lty = 2,
        color = "black"
      ) +
      ggtitle(paste(PairwiseGroups_title, "padj 0.01")) +
      xlim(-xbound, xbound) +
      xlab("Average log2 fold change") +
      ylab("-log10 adjusted p-value") +
      theme_bw(base_size = 12) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        panel.grid.minor = element_blank()
      )
    
    print(gg)
    
    # unique volcano filename per comparison
    volcano_file <- make_path(dir = dir_plots,
                              filename = paste0("VolcanoPlot_",
                                                PairwiseGroups_title, "_padj001"))

    ggsave(paste0(volcano_file, ".pdf"), plot = gg, width = 10, height = 6)
    ggsave(paste0(volcano_file, ".png"), plot = gg, width = 10, height = 6)
    ggsave(paste0(volcano_file, ".svg"), plot = gg, width = 10, height = 6)

  }
  
}
```



## Significance

    **Volcano plot threshold is set to padj 0.05**
    
    **Significant genes of padj 0.01 are used for downstream analysis of GO**

    **p-value of 0.05 was chosen for GSEA analysis**


***


# GSEA for human CD4 CTL from IEL_CD4

## Setup

```{r}
Idents(int.obj.cell) <- "HTO_classification"
DefaultAssay(int.obj.cell) <- "SCT"
```


## Term2Gene
```{r}
# Build TERM2GENE
term2gene_custom <- data.frame(
  term = rep("custom_gene_set", length(gene_list)),
  gene = gene_list,
  stringsAsFactors = FALSE
)

head(term2gene_custom)
```

## GSEA with human list

```{r}
  for (i in seq_along(DEgroups_HTOclasses)) {
    
    group1 <- DEgroups_HTOclasses[[i]][[1]]  # KO
    group2 <- DEgroups_HTOclasses[[i]][[2]]  # WT
    
    group1_collapsed <- paste(group1, collapse = " ")
    group2_collapsed <- paste(group2, collapse = " ")
    
    PairwiseGroups_title <- paste0(group1_collapsed, "_vs_", group2_collapsed)
    
    writeLines(paste0("Pairwise Group Comparison ", i, ":"))
    writeLines(paste0("Group1: ", group1_collapsed))
    writeLines("versus")
    writeLines(paste0("Group2: ", group2_collapsed))
    
    # Differential expression (no thresholds: full ranking)
    markers <- FindMarkers(
      int.obj.cell,
      ident.1        = group1,
      ident.2        = group2,
      assay          = "SCT",
      logfc.threshold = 0,
      min.pct         = 0
    )
    
    
    
    
    # Drop NAs in avg_log2FC
    markers <- markers[!is.na(markers$avg_log2FC), , drop = FALSE]
    if (nrow(markers) == 0) {
      message("No markers returned for ", PairwiseGroups_title, " – skipping GSEA.")
      next
    }
    
    # Prepare ranked gene list for GSEA
    ranked_genes <- markers$avg_log2FC
    names(ranked_genes) <- rownames(markers)
    
    # Sort decreasing, remove duplicates
    ranked_genes <- sort(ranked_genes, decreasing = TRUE)
    ranked_genes <- ranked_genes[!duplicated(names(ranked_genes))]
    
    # check overlap between custom gene set and ranked genes
    overlap_genes <- intersect(names(ranked_genes), gene_list)
    if (length(overlap_genes) < 5) {
      message("Less than 5 overlapping genes between ranked list and custom gene set for ",
              PairwiseGroups_title, " – GSEA likely uninformative")
      next
    }
    
    
    ####
    # Run GSEA for the single custom term
    gsea_results <- GSEA(
      geneList = ranked_genes,
      TERM2GENE = term2gene_custom,
      pvalueCutoff = 0.05,
      verbose       = TRUE
    )
    gsea_results_table <- as.data.frame(gsea_results)
    
    # Save GSEA result table
    output_file <- make_path(dir = dir_tables, filename = paste0("GSEA_results_pval005_", PairwiseGroups_title, ".csv"))
    
    write.csv(gsea_results_table, output_file, row.names = FALSE)
    writeLines(paste0("GSEA results saved to: ", output_file))
    
    # Extract NES and FDR
    # Normalized Enrichment Score = NES
    NES <- gsea_results_table$NES[1] # "custom_gene_set" is the first row
    padj <- gsea_results_table$p.adjust[1] # padj correspond to FDR
    
    # Plot GSEA enrichment scores with annotations
    enrichment_plot <- gseaplot2(
      gsea_results,
      geneSetID = "custom_gene_set",
      title = paste("GSEA Enrichment:", PairwiseGroups_title, "NES:", round(NES, 2), "  | adj.p-val = ", round(padj, 5)),
      base_size = 12
    )
    
    print(enrichment_plot)
    
    plot_file <- make_path(dir = dir_plots, filename = paste0("GSEA_enrichment_plot_", PairwiseGroups_title))
    
    ggsave(paste0(plot_file, ".pdf"), plot = enrichment_plot, width = 8, height = 6)
    ggsave(paste0(plot_file, ".png"), plot = enrichment_plot, width = 8, height = 6)
    ggsave(paste0(plot_file, ".svg"), plot = enrichment_plot, width = 8, height = 6)
    
    writeLines(paste0("GSEA plot saved to: ", plot_file))
    
    
    # Filter gene SETs with positive NES (or ES, if available)
    filtered_genes <- gsea_results_table[gsea_results_table$NES > 0, ]
    
    # Check if data exists before exporting
    if (nrow(filtered_genes) > 0) {
      # Define output filename
      excel_file <- make_path(dir= dir_tables, filename = paste0("GSEA_positive_NES_GeneSets_", PairwiseGroups_title, ".xlsx"))
      
      # Write to Excel
      write.xlsx(filtered_genes, excel_file, row.names = FALSE)
      writeLines(paste0("Excel file saved: ", excel_file))
    } else {
      writeLines("No genes found with NES > 0.")
    }
    
  }


```


## Save RDS
```{r}
obj_file <- file.path(dir_obj, paste0("08_", TISSUE, "_", CELL, "_IntestinalSeurat_reclustered.rds"))

saveRDS(int.obj.cell, obj_file)

```

